### 2009-10-29 doc/papers/2010-icgl-hongkong/paper.tex
Přidány články. LREC 2008 už je uzavřený, naopak ICGL 2010 teprve bude vznikat, tj. teď je tam kopie nějakého úplně jiného článku.

% !TEX TS-program = xelatex
% !TEX encoding = UTF-8 Unicode
% File acl-ijcnlp2009.tex
%
% Contact  jshin@csie.ncnu.edu.tw
%%
%% Based on the style files for EACL-2009 and IJCNLP-2008...
%% Based on the style files for EACL 2006 by 
%%e.agirre@ehu.es or Sergi.Balari@uab.es
%% and that of ACL 08 by Joakim Nivre and Noah Smith

\documentclass[11pt]{article}
\usepackage{acl-ijcnlp2009}
\usepackage[	
   pdfdisplaydoctitle, breaklinks, colorlinks, linkcolor=black, citecolor=black, filecolor=black, urlcolor=black, 
   backref, hyperfootnotes]{hyperref} % backref a modre URL asi nakonec zrusime 
%\usepackage{times}
\usepackage{url}
\usepackage{amsmath}
\usepackage{color} %pro korektury
\usepackage{paralist} % for better itemize and enumerate

% a footer required for the first page
\usepackage{fancyhdr}
\fancyhead{} % clear all header fields
\renewcommand{\headrulewidth}{0pt}
\fancyfoot[C]{Proceedings of ICON-2009: 7th International Conference on Natural Language Processing, Macmillan Publishers, India. Also accessible from http://ltrc.iiit.ac.in/proceedings/ICON-2009}

% xelatex
\usepackage{fontspec, xunicode, xltxtra}
\defaultfontfeatures{Mapping=tex-text}
\setmainfont{Times New Roman}
\setmonofont[Scale=MatchLowercase]{Luxi Mono}
\setmathsf{Lohit Hindi}%\XXX
%\newfontinstance\hi[Script=Devanagari]{Lohit Hindi}
\newfontinstance\hifont[Script=Devanagari]{Code2000}
\newfontinstance\bnfont[Script=Bengali]{Code2000}
\newfontinstance\tefont[Script=Telugu]{Code2000}
\newfontinstance\translitfont{Gentium}
\newcommand{\hi}[1]{{\hifont #1}}
\newcommand{\bn}[1]{{\bnfont #1}}
\newcommand{\te}[1]{{\tefont #1}}
\newcommand{\translit}[1]{{\translitfont \textit{(#1)}}}

% natbib
\usepackage{natbib}
\bibliographystyle{plainnat}
\bibpunct{(}{)}{;}{a}{,}{,}

% our defs
\def\perscite#1{\citet{#1}}  
\def\parcite#1{\citep{#1}} 
%ps: Did you mean \citep and \citet (in-parentheses and textual reference)? There is even more, see `texdoc natbib`.
\def\Sref#1{Section~\ref{#1}}
\def\Tref#1{Table~\ref{#1}}
\def\Fref#1{Figure~\ref{#1}}
\newcommand{\red}[1]{\textcolor{red}{#1}} % komentare (TODO)
\newcommand{\XXX}{\textcolor{red}{XXX }} % komentare (TODO)

\def\microsection#1{{\bf #1.}}

\title{Maximum Spanning Malt: Hiring World's Leading Dependency Parsers to Plant Indian Trees%
% Tohle nechat zakomentované, je to jen tahák, jak udělat acknowledgement grantu při nedostatku místa. Jinak ale mám momentálně na konci opravdovou sekci Acknowledgements.
%\thanks{ \hspace{.6em}The research has been supported by the grant 
%MSM0021620838 (Czech Ministry of Education).}
}

\author{Daniel Zeman\\
Univerzita Karlova v Praze, Ústav formální a aplikované lingvistiky\\
Malostranské náměstí 25, CZ-11800, Praha, Czechia\\ 
\texttt{zeman@ufal.mff.cuni.cz}
}

%\title{Instructions for ACL-IJCNLP 2009 Proceedings}
%
%\author{First Author\\
%  Affiliation / Address line 1\\
%  Affiliation / Address line 2\\
%  {\tt email@domain}  \And
%  Second Author\\
%  Affiliation / Address line 1\\
%  Affiliation / Address line 2\\
%  {\tt  email@domain}}

\date{}

\begin{document}
\maketitle
\thispagestyle{fancy}

\begin{abstract}
We present our system used for participation in the ICON 2009 NLP Tools Contest: dependency parsing of Hindi, Bangla and Telugu. The system consists of three existing, freely available dependency parsers, two of which (MST and Malt) have been known to produce state-of-the-art structures on data sets for other languages. Various settings of the parsers are explored in order to adjust them for the three Indian languages, and a voting approach is used to combine them into a superparser. Since there is nothing novel about the approach used, substantial part of the paper is devoted to the analysis of errors the system makes on the given data sets.
\end{abstract}

\section{Introduction}
\label{sec:intro}

Dependency parsing, i.e. sentence analysis that outputs tree of word-on-word dependencies (as opposed to constituent trees of context-free derivations), gained growing attention and popularity recently. There are data-driven dependency parsers that can be trained on syntactically annotated corpora (treebanks) and new, previously unseen material can be parsed very efficiently \citep{nivre:2009:ACLIJCNLP}.

Most of the successful parsers employ discriminative learning techniques to sort out vast sets of potentially useful features observed in the input text. Thus, for every new training treebank, smart feature engineering is the key to getting the most out of the existing parsers, regardless how well they performed on other data sets and languages. Now that there are new treebanks available for two Indo-Aryan and one Dravidian language, we took three existing dependency parsers and explored the possibilities of tuning them for the new training data. Both parser configuration and data preprocessing are relevant approaches to the tuning. In addition, we used parser combination to further improve the results.

Throughout the paper we focus mainly on the unlabeled attachment score. Although the parsers produce labeled dependencies, we do not optimize the system towards label accuracy.

The rest of the paper is organized as follows: In \Sref{sec:system}, we describe the respective parsers and the combined parsing system. In \Sref{sec:experiments}, we report on the experiments we performed, discuss various results on the development set and analyze the errors. In \Sref{sec:evaluation} we present the official results on the test data. We conclude by summarizing the best configuration we were able to find, and future implications.

\section{System Description}
\label{sec:system}

Several good trainable dependency parsers have emerged during the past five years. The CoNLL-X \citep{buchholz-marsi:2006:CoNLL-X} and CoNLL 2007 \citep{nivre-EtAl:2007:EMNLP-CoNLL2007} shared tasks in multilingual dependency parsing have greatly contributed to the development of the parsers. Some of the parsers are now freely available on the web, some are even open-source. We selected three of the publicly available parsers for our experiments:

%\microsection{MST Parser}
\subsection{MST Parser}
\label{sec:mst}
The Maximum Spanning Tree (MST) parser \citep{mst} views the sentence as an directed complete graph with edges weighted by a feature scoring function. It finds for the graph the spanning tree that maximizes the weights of the edges. A multi-class classification algorithm called MIRA is used to compute the scoring function.

MST Parser achieved the best unlabeled attachment scores (UAS) for 9 out of the 13 languages of CoNLL-X, and second best scores in two others. Parsing is fast but training the parser takes many hours on large treebanks. On small data however, multiple quick experiments with different settings are still doable. The parser is implemented in Java and freely available for download.\footnote{\url{http://sourceforge.net/projects/mstparser/}}

%\microsection{Malt Parser}
\subsection{Malt Parser}
\label{sec:malt}
The Malt Parser  \citep{malt} is a deterministic shift-reduce parser where input words can be either put to the stack or taken from the stack and combined to form a dependency. The decision which operation to perform is made by an oracle based on various features of the words in the input buffer and the stack. The default machine learning algorithm used to train the oracle is a sort of SVN (support vector machine) classifier \citep{svm}.

Malt Parser has participated in both CoNLL-X and CoNLL 2007 shared tasks, and although it achieved the best UAS in three languages only, it usually scored among the five best parsers, sometimes with statistically insignificant difference from the winner. Malt Parser is really fast and its new Java implementation is open-source, freely available for download.\footnote{\url{http://maltparser.org/}}

%\microsection{DZ Parser}
\subsection{DZ Parser}
\label{sec:dz}
In order to combine the two above parsers, we needed a third parser. We picked DZ Parser \citep{dzparser}, which is also reasonably fast and freely available.\footnote{\url{http://ufal.mff.cuni.cz/~zeman/projekty/parser/}} Although its accuracy, if compared to MST or Malt, is worse by a wide margin, this parser proved useful because its only role was to help to form a majority whenever MST and Malt disagreed.

DZ Parser builds a model of bigrams of words that occur together in a dependency; most of the time, words are identified by their part of speech tags and morphological features. The parser was originally developed for Czech but it can be re-trained for any other language.\footnote{Of course there are other dependency parsers that successfully participated in the CoNLL shared tasks and are available for download. One alternative worth mentioning is the ISBN Parser \citep{titov-henderson:2007:EMNLP-CoNLL2007} at \url{http://flake.cs.uiuc.edu/~titov/}.}

\subsection{Voting Superparser}
\label{sec:voting}
The three parsers are combined using a simple weighted-voting approach similar to \citet{biblio:ZeZaImprovingParsing2005}, except that the output is guaranteed to be cycle-free. We start by evaluating every parser separately on the development data. The UAS of each parser is subsequently used as the weight of that parser's vote. Dependencies are parent-child relations, and for every node there are up to three candidates for its parent (if all three parsers disagree). Candidates get weighted votes -- e.g., if parsers with weights $w_1 = 0.8$ and $w_2 = 0.7$ agree on the candidate, the candidate gets 1.5 votes. Since we have only three parsers, in practice this means that the candidate of the best parser looses only if 1. the other two parsers agree on someone else, or 2. if attaching the child to this candidate would create a cycle.

The tree is constructed from the root down. We repeatedly add nodes whose winning parent candidates are already in the tree. If none of the remaining nodes meet this condition, we have to break a cycle. We do so by examining all unattached nodes. At each node we notice the votes of its current winning parent. Then we remove the least-scoring winner and go on with adding nodes until all nodes are attached or there is another cycle to break.

\section{Experiments}
\label{sec:experiments}

The final test data are blind, any error analysis is therefore impossible. That is why all scores given in this section were measured on the development data. All three treebanks follow the same annotation scheme and each of them is available in two flavors:

\begin{compactitem}
\item \textit{nomorph} variety contains word forms, chunk labels, dependency links and dependency labels
\item \textit{morph} variety is augmented by automatically assigned lemmas, part of speech tags and values of morphological features (gender, number, person, case, postposition and tam -- tense+aspect+modality)
\end{compactitem}

\subsection{Morphology}
\label{sec:morphology}

\Tref{tab:baseline} shows baseline results on the \textit{nomorph} data. Both MST and Malt parsers were invoked in projective mode, Malt with the default Nivre arc-eager algorithm.

\begin{table}[ht]
\begin{centering}
%\small
\begin{tabular}{l|l|l|l|l}
& \textbf{MST} & \textbf{Malt} & \textbf{DZ} & \textbf{Vote} \\
\hline
hi & 80.32 & 81.84 & 62.00 & \textbf{82.48}\\
bn & 82.00 & \textbf{84.71} & 71.02 & 83.11\\
te & 77.63 & \textbf{80.89} & 70.52 & 80.59\\
\end{tabular}
\caption{Baseline UAS of the four parsers on \textit{nomorph} development data. Language codes follow ISO 639: hi = Hindi, bn = Bangla, te = Telugu.}
\label{tab:baseline}
\end{centering}
\end{table}

There are several ways how to use the additional information from the \textit{morph} data. The easiest way, exploitable by all three parsers, is to combine the chunk label, the POS tag and the features into one tag string. The results (not presented here) are very poor. Although there may be tagging errors, the most likely cause is data sparseness. In \Tref{tab:corpus} we illustrate this by showing the numbers of unique values in the various attributes of treebank words.

\begin{table}[ht]
\begin{centering}
%\small
\begin{tabular}{l|r|r|r|r|r|r}
& \textbf{occ} & \textbf{frm} & \textbf{lem} & \textbf{cl} & \textbf{pos} & \textbf{feat} \\
\hline
hi & 13779 & 3973 & 3134 & 10 & 33 & 714\\
bn & 6449 & 2997 & 2336 & 14 & 30 & 367\\
te & 5494 & 2462 & 1403 & 12 & 31 & 453\\
\end{tabular}
\caption{Size of the training corpora: occ -- word occurrences, frm -- distinct forms, lem -- lemmas, cl -- chunk labels, pos -- part of speech tags, feat -- feature value combinations.}
\label{tab:corpus}
\end{centering}
\end{table}

To fight sparseness, we could either restrict the tag to selected information, or split the information into multiple features learnt separately, or both. We first restricted the tag string to selected information. Parsing on other treebanks showed that POS with case is especially useful \citep{dzparser}. The other feature we selected is called \textit{vibhakti}, which partially corresponds to case suffix and partially to postposition.\footnote{The Indian treebanks at hand are unusual in that nodes do not always map to words. They represent chunks, with function words such as postpositions hidden in node attributes.}

\Tref{tab:poscasevib} presents the results of restricting the tag string to POS+case+vibhakti. This move especially improves the MST Parser, which now outperforms Malt on Hindi and Bangla. Malt improves on Hindi but drops behind on Bangla and Telugu. DZ Parser also improves on Hindi and deteriorates elsewhere but there is an interesting observation: even though its own scores are worse, the worsened output actually improves the voting results (provided the configurations of MST and Malt are fixed). So it seems that the newly introduced errors are less important (because taken care of by the more powerful parsers) while some difficult parts of the data are now covered better.

\begin{table}[ht]
\begin{centering}
%\small
\begin{tabular}{l|l|l|l|l}
& \textbf{MST} & \textbf{Malt} & \textbf{DZ} & \textbf{Vote}\\
\hline
hi & 86.16 & 82.96 & 75.12 & \XXX 86.64\\
bn & 85.70 & 80.02 & 54.38 & \XXX 78.79\\
te & 79.85 & 78.37 & 45.78 & \XXX 78.81\\
\end{tabular}
\caption{UAS on refined \textit{morph} data (POS tag, case and vibhakti concatenated). \XXX voting se pocitalo, kdyz jeste MST pracoval jen s nomorph, takze to musime prepocitat}
\label{tab:poscasevib}
\end{centering}
% Když se ke značce POS přilepí pád, ale ne vibhakti, a s tím se přetrénuje Malt parser (ostatní parsery jsme nechali beze změny, tj. MST byl natrénovaný na nomorph a DZ naopak i na vibhakti), tak Malt sám má na všech třech jazycích pořád stejné výsledky (jako v této tabulce), ale výsledky Vote se na hindštině zhorší a na zbývajících dvou jazycích zlepší: hi=85.36, bn=83.35, te=80.89.
\end{table}

Finally, ve returned to \textit{nomorph} with Malt on Bangla and Telugu, where the POS+case+vibhakti tags did not help. The results are given in \Tref{tab:posmix2}. This is also the configuration we used to parse the test set for the official evaluation round 1.

\begin{table}[ht]
\begin{centering}
%\small
\begin{tabular}{l|l|l|l|l}
& \textbf{MST} & \textbf{Malt} & \textbf{DZ} & \textbf{Vote}\\
\hline
hi & 86.16 & 85.84 & 75.12 & \textbf{87.12}\\
bn & 85.70 & 84.71 & 54.38 & \textbf{86.19}\\
te & 79.85 & 80.89 & 45.78 & \textbf{82.37}\\
\end{tabular}
\caption{UAS on mixed data: MST and DZ use POS+case+vibhakti for all languages, Malt uses that for Hindi and POS only elsewhere. MST is now the best parser for hi and bn, which changes the voting weights.}
\label{tab:posmix2}
\end{centering}
\end{table}

%\begin{compactitem}
%\item \hi{स्टैंडर्डज} \textit{(sṭaiṁḍarḍaja)} \translit{(sṭaiṁḍarḍaja)}
%\item \hi{स्टैंडर्डस} \textit{(sṭaiṁḍarḍasa)}
%\item \hi{स्टैंडर्ड्स} \textit{(sṭaiṁḍarḍsa)}
%\item \bn{সরিয়ে} Tohle je bengálsky.
%\item \te{వాడుతున్నాం} Tohle je telugsky.
%\end{compactitem}

\subsection{Nonprojectivity}
\label{sec:nonprojectivity}

Nonprojectivity is a property of the dependency structure and the word order \citep{neproj} that makes parsing more difficult. All three parsers can produce nonprojective structures and all three treebanks are nonprojective. However, except for Hindi, the proportion of nonprojective dependencies is so small that one can hardly imagine that running the parsers in nonprojective mode would bring any improvement. A quick experiment with Malt Parser switched to the nonprojective Stack Eager algorithm revealed that it actually hurts the results even for Hindi.

\begin{table}[ht]
\begin{centering}
%\small
\begin{tabular}{l|ll}
& \textbf{Edges} & \textbf{Sentences} \\
\hline
hi & 01.83 & 13.93\\
bn & 00.96 & 05.49\\
te & 00.45 & 01.31\\
\end{tabular}
\caption{Percentage of nonprojective dependencies, and of sentences containing at least one nonprojectivity.}
\label{tab:nonprojectivity}
\end{centering}
\end{table}

\subsection{Error Patterns}
\label{sec:errors}

The accuracy of the dependencies is relatively high and it is difficult to trace repetitive error patterns. In Hindi, many wrong attachments seem to be long-distance, and verbs, conjunctions, root and NULL nodes are frequently involved. Frequent words should perhaps be available to the parsers as parts of tag strings: for instance, Hindi \hi{कि} \translit{ki} ``that'' or \hi{तो} \translit{to} are wrongly attached because the parser only sees the general CC tag. On a similar note, problems with coordination, also observed e.g. by \citet{dzparser}, occur here, too: \hi{भाई और भाभी} \translit{bhāī aura bhābhī} ``brother and his wife'' is correctly recognized as coordination rooted by the conjunction \hi{और}, however, the conjunction node lacks the information about its noun children and fails to attach as the subject of the verb.

The tag string should contain both the chunk label and the POS tag. So far we wrongly assumed that POS always determins the chunk label. It is often so but not always, as exemplified in the Bangla chunk sequence \bn{তবে সুদীপ ওকে একদিন আড়ালে ডেকে বলেছিল কৌতূহল দেখালে তুমি উঁচুতে উঠতে অনিমেষ} \translit{tabe sudīpa oke ekadina āṛāle ḍeke balechila kautūhala dekhāle tumi um̃cute uṭhate animeša}. The words \bn{ডেকে} and \bn{দেখালে} are tagged VGNF|VM while \bn{বলেছিল} and \bn{উঠতে} are VGF|VM. The parser gets them wrong and it could be caused by it seeing only VM in the tag.

In Telugu, extraordinal number of sentences follow the SOV order so strongly that the last node (verb) is almost always attached to the root and most other nodes are attached directly to the last node. An example chunk sequence where this rule would lead to 100~\% accuracy follows: \te{రాష్ట్రంలొ రంగారెడ్డి మెదక్ నిజామాబాద్ జిల్లాలలొ పంటను గొప్పొ పండిస్తున్నారు} \translit{rāšṭraṁlo raṁgāreḍḍi medak nijāmābād jillālalo paṁṭanu goppo paṁḍistunnāru}. In the light of such examples it seems reasonable to provide the parsers with an additional feature telling whether a particular dependency observes the ``naïve Telugu'' structure. Note however that this will not help with the other two languages. While 73.75~\% of Telugu dependncies follow this rule, it is only 39.52~\% in Bangla and 35.71~\% in Hindi.

\subsection{\XXX To Do}
\begin{compactitem}
\item \XXX Parser unique errors (oracle accuracy)
\item \XXX How many times a candidate lost due to cycle prevention, and how many times this introduced an error?
\item \XXX Learning curve
\item \XXX Spočítat, kolik je ve kterém jazyce uzlů NULL.
\item \XXX znacka chunku, priklad z bengalstiny
\item \XXX casta slova, priklad z hindstiny
\item \XXX koordinace, priklad z hindstiny
\end{compactitem}

\section{Official Evaluation}
\label{sec:evaluation}

Finally, we present the official evaluation of our voting superparser, as measured by the organizers on the test data. For this purpose, the parsing system has been retrained on both the training data and the development data. The results are shown in \Tref{tab:evaluation}.

\begin{table}[ht]
\begin{centering}
\small
\begin{tabular}{l|l|l|l}
& \textbf{UAS} & \textbf{LAA} & \textbf{LAS} \\
\hline
hi & 88.58 (3:90.31) & 72.66 (4:76.38) & 68.60 (4:74.48)\\
bn & 86.06 (4:90.32) & 71.28 (4:81.27) & 66.70 (5:79.81)\\
te & 80.27 (4:86.28) & 54.20 (4:61.58) & 49.91 (4:60.55)\\
\end{tabular}
\caption{Official scores on the test data: unlabeled attachment score (UAS), label assignment accuracy (LAA) and labeled attachment score (LAS). The numbers in parentheses are the rank of our system and the score of the best system w.r.t. the given metric.}
\label{tab:evaluation}
\end{centering}
\end{table}

\section{Related and Future Work}
\label{sec:related}

There is a large body of work on parser combination. A summary can be found in \citet{nivre-mcdonald:2008:ACLMain}, whose approach is also related to ours w.r.t. the selection of parsers. However, their feature-based integration of MST and Malt parsers is much more sophisticated than our lightweight voting. Further improvement of accuracy can be expected if MST-Malt integration is applied to the Indian treebanks.

Future work, at the time of writing, includes experiments that can be run before the evaluation round 2, and thus their results will appear in the final version of this paper. We definitely intend to test other algorithms of the Malt parser, as well as to reconfigure its feature pool and let it work with POS, case and vibhakti separately.

Labeling of the dependencies is another problem that deserves more attention. We have concentrated on the unlabeled attachment score so far and for the sake of the official evaluation, we simply pushed the MST labels through. A separate postprocessing classifier would probably produce better results.

\section{Conclusion}
\label{sec:concl}

We have described our system of voting parsers, as applied to the ICON 2009 NLP Tools Contest task. We showed that case and vibhakti are important features at least for parsing Hindi while their usability in Bangla and Telugu is limited by data sparseness. We also discussed several error patterns that could lead to further improvements of the parsing system in future.

\section*{Acknowledgements}

We are enormously grateful to the developers of MST and Malt parsers for making their software available to the research community.
The research has been supported by the grant 
MSM0021620838 (Czech Ministry of Education).

\begin{small}
\bibliography{paper}
\end{small}

\end{document}

### 2009-10-30 doc/papers/2010-icgl-hongkong/paper.tex
Zárodek článku ICGL.

@@
 
-% a footer required for the first page
+% a footer required for the first page (ICON but not ICGL)
 \usepackage{fancyhdr}
@@
 \renewcommand{\headrulewidth}{0pt}
-\fancyfoot[C]{Proceedings of ICON-2009: 7th International Conference on Natural Language Processing, Macmillan Publishers, India. Also accessible from http://ltrc.iiit.ac.in/proceedings/ICON-2009}
+\fancyfoot[C]{
+%Proceedings of ICON-2009: 7th International Conference on Natural Language Processing, Macmillan Publishers, India. Also accessible from http://ltrc.iiit.ac.in/proceedings/ICON-2009
+}
 
@@
 
-\title{Maximum Spanning Malt: Hiring World's Leading Dependency Parsers to Plant Indian Trees%
+\title{Hard Problems of Tagset Conversion%
 % Tohle nechat zakomentované, je to jen tahák, jak udělat acknowledgement grantu při nedostatku místa. Jinak ale mám momentálně na konci opravdovou sekci Acknowledgements.
@@
 
-\author{Daniel Zeman\\
-Univerzita Karlova v Praze, Ústav formální a aplikované lingvistiky\\
-Malostranské náměstí 25, CZ-11800, Praha, Czechia\\ 
-\texttt{zeman@ufal.mff.cuni.cz}
+% Až to přestane být anonymní, tak také odkomentovat Acknowledgements a autoreference XXXX.
+\author{%Daniel Zeman\\
+%Univerzita Karlova v Praze, Ústav formální a aplikované lingvistiky\\
+%Malostranské náměstí 25, CZ-11800, Praha, Czechia\\ 
+%\texttt{zeman@ufal.mff.cuni.cz}
 }
@@
 \begin{abstract}
-We present our system used for participation in the ICON 2009 NLP Tools Contest: dependency parsing of Hindi, Bangla and Telugu.
-The system consists of three existing, freely available dependency parsers, two of which (MST and Malt) have been known to produce state-of-the-art structures on data sets for other languages.
-Various settings of the parsers are explored in order to adjust them for the three Indian languages, and a voting approach is used to combine them into a superparser.
-Since there is nothing novel about the approach used, substantial part of the paper is devoted to the analysis of errors the system makes on the given data sets.
+Part-of-speech or morphological tags are important means of annotation in a vast number of corpora.
+However, different sets of tags are used in different corpora, even for the same language.
+Tagset conversion is difficult, and solutions tend to be tailored to a particular pair of tagsets.
+We propose a universal approach that makes the conversion tools reusable.
+\XXX We also provide an indirect evaluation in the context of a parsing task.
 \end{abstract}
@@
 
-Dependency parsing, i.e.
-sentence analysis that outputs tree of word-on-word dependencies (as opposed to constituent trees of context-free derivations), gained growing attention and popularity recently.
-There are data-driven dependency parsers that can be trained on syntactically annotated corpora (treebanks) and new, previously unseen material can be parsed very efficiently \citep{nivre:2009:ACLIJCNLP}.
+Most annotated corpora use various types of tags to encode additional information on words.
+In some cases this information is merely the part of speech (“noun”, “verb” etc.—hence the term \textit{part-of-speech} or \textit{POS tags}).
+In many cases, however, the string of characters comprising the tag is a compressed representation of a feature-value structure.
+Most of the features encoded this way are morphosyntactic (e.g.
+“gender = masculine”, “number = singular”), hence the term \textit{morphological tags}.
 
-Most of the successful parsers employ discriminative learning techniques to sort out vast sets of potentially useful features observed in the input text.
-Thus, for every new training treebank, smart feature engineering is the key to getting the most out of the existing parsers, regardless how well they performed on other data sets and languages.
-Now that there are new treebanks available for two Indo-Aryan and one Dravidian language, we took three existing dependency parsers and explored the possibilities of tuning them for the new training data.
-Both parser configuration and data preprocessing are relevant approaches to the tuning.
-In addition, we used parser combination to further improve the results.
+Unfortunately, it is very rare to see two corpora sharing a common set of tags.
+Language differences are only partially responsible—it is the corpus designers, their diverse views, theories and intended uses of the corpora, what matters most.
+Even two corpora of the same language may define two completely incompatible tagsets.
 
-Throughout the paper we focus mainly on the unlabeled attachment score. Although the parsers produce labeled dependencies, we do not optimize the system towards label accuracy.
+Such diversity proves disadvantageous for both human users and NLP software.
+A human user (linguist) typically wants to submit queries such as “show me all occurrences of a noun in plural, preceded by a preposition”.
+Tags however rarely contain statements like “number = plural” literally.
+That would be prohibitively space-consuming.
+Instead we have to know that e.g.
+the fourth character of the tag being “P” means “plural”.
+For instance, the tag \texttt{NNIS7-{}-{}-{}-{}-A-{}-{}-{}-}\footnote{This example is taken from the Prague Dependency Treebank \citep{pdt}.} may read as “part of speech = noun, detailed part of speech = common noun, gender = masculine inanimate, number = singular, case = 7th (instrumental), negativeness = affirmative”.
+To work with the corpus efficiently, a linguist either needs to interpret the tags using specialized software, or to memorize the particular tag scheme.
+Obviously, if the same linguist has to switch to a different corpus, he/she must memorize more schemes or replace the tag interpretation software.
 
-The rest of the paper is organized as follows: In \Sref{sec:system}, we describe the respective parsers and the combined parsing system.
-In \Sref{sec:experiments}, we report on the experiments we performed, discuss various results on the development set and analyze the errors.
-In \Sref{sec:evaluation} we present the official results on the test data.
-We conclude by summarizing the best configuration we were able to find, and future implications.
+Similarly, various NLP tools may depend on particular tagsets.
+While some tools indeed treat tags as atomic strings, others could exploit the tag structure to dig more information about the word—no matter whether they use the features in machine learning, or in human-designed rules.
+If the tagset changes, manual rules become useless and statistical models have to be retrained at least; even that may not be possible in case the training procedure works with selected subsets of the feature pool.
+Applicability of NLP software to multiple corpora is exactly the reason why one would want to convert tags from one tagset to another.
 
-\section{System Description}
-\label{sec:system}
+For many tagset pairs, designing the conversion procedure is not easy.
+On one hand, there are rare tagsets (e.g.
+MULTEXT-EAST, \citet{multext-east}) fitting at the same time languages as distant as Czech and Estonian; on the other hand, tagsets of two closely related languages (e.g.
+Danish and Swedish) or even two tagsets of the same language may differ substantially (for instance, the Mamba tagset of Swedish \citep{talbanken} contains detailed classification of auxiliary verbs and punctuation but lacks features like number, mood, tense etc.; this is in sharp contrast to another Swedish tagset, Parole \citep{parolesv}, which in turn is not compatible with the Danish Parole \citep{paroleda} tagset (the former classifies participles as verb forms, the latter as adjective forms; the former has separate tags for numerals, the latter classifies both cardinal and ordinal numbers as adjectives; etc.)
 
-Several good trainable dependency parsers have emerged during the past five years.
-The CoNLL-X \citep{buchholz-marsi:2006:CoNLL-X} and CoNLL 2007 \citep{nivre-EtAl:2007:EMNLP-CoNLL2007} shared tasks in multilingual dependency parsing have greatly contributed to the development of the parsers.
-Some of the parsers are now freely available on the web, some are even open-source.
-We selected three of the publicly available parsers for our experiments:
+From the above said it follows that the typical tag conversion is an information-losing process.
+Though it is often desirable to perform it anyway and preserve as much information as possible.
+Creation of a conversion procedure between two tagsets requires hours of tedious work, consisting mostly of reading the tagging guidelines and translating them into a programming language.
+A universal description, to which all tagsets map, could make this process easier, and its results reusable.
+One attempt to find such description and deploy it in the conversion task is DZ Interset \citep{biblio:ZeReusableTagset2008}.
+In the present paper we discuss the development of the universal description and focus on selected hard problems that arise when comparing various existing tagsets.
 
-%\microsection{MST Parser}
-\subsection{MST Parser}
-\label{sec:mst}
-The Maximum Spanning Tree (MST) parser \citep{mst} views the sentence as an directed complete graph with edges weighted by a feature scoring function.
-It finds for the graph the spanning tree that maximizes the weights of the edges.
-A multi-class classification algorithm called MIRA is used to compute the scoring function.
+The rest of the paper is organized as follows: In \XXX we describe Interset and how it works [\XXX including encoding algorithm] [\XXX including applications: Petr Pořízka, Saša Rosen, můj parsing].
+In \XXX we describe our universal set of features.
+Then, \XXX lists decisions that are difficult w.r.t.
+universality, and propose solutions.
+Finally, we demonstrate the implications on real tagsets, and provide illustrative statistics.
 
-MST Parser achieved the best unlabeled attachment scores (UAS) for 9 out of the 13 languages of CoNLL-X, and second best scores in two others.
-Parsing is fast but training the parser takes many hours on large treebanks.
-On small data however, multiple quick experiments with different settings are still doable.
-The parser is implemented in Java and freely available for download.\footnote{\url{http://sourceforge.net/projects/mstparser/}}
-
-%\microsection{Malt Parser}
-\subsection{Malt Parser}
-\label{sec:malt}
-The Malt Parser  \citep{malt} is a deterministic shift-reduce parser where input words can be either put to the stack or taken from the stack and combined to form a dependency.
-The decision which operation to perform is made by an oracle based on various features of the words in the input buffer and the stack.
-The default machine learning algorithm used to train the oracle is a sort of SVN (support vector machine) classifier \citep{svm}.
-
-Malt Parser has participated in both CoNLL-X and CoNLL 2007 shared tasks, and although it achieved the best UAS in three languages only, it usually scored among the five best parsers, sometimes with statistically insignificant difference from the winner.
-Malt Parser is really fast and its new Java implementation is open-source, freely available for download.\footnote{\url{http://maltparser.org/}}
-
-%\microsection{DZ Parser}
-\subsection{DZ Parser}
-\label{sec:dz}
-In order to combine the two above parsers, we needed a third parser.
-We picked DZ Parser \citep{dzparser}, which is also reasonably fast and freely available.\footnote{\url{http://ufal.mff.cuni.cz/~zeman/projekty/parser/}} Although its accuracy, if compared to MST or Malt, is worse by a wide margin, this parser proved useful because its only role was to help to form a majority whenever MST and Malt disagreed.
-
-DZ Parser builds a model of bigrams of words that occur together in a dependency; most of the time, words are identified by their part of speech tags and morphological features.
-The parser was originally developed for Czech but it can be re-trained for any other language.\footnote{Of course there are other dependency parsers that successfully participated in the CoNLL shared tasks and are available for download.
-One alternative worth mentioning is the ISBN Parser \citep{titov-henderson:2007:EMNLP-CoNLL2007} at \url{http://flake.cs.uiuc.edu/~titov/}.}
-
-\subsection{Voting Superparser}
-\label{sec:voting}
-The three parsers are combined using a simple weighted-voting approach similar to \citet{biblio:ZeZaImprovingParsing2005}, except that the output is guaranteed to be cycle-free.
-We start by evaluating every parser separately on the development data.
-The UAS of each parser is subsequently used as the weight of that parser's vote.
-Dependencies are parent-child relations, and for every node there are up to three candidates for its parent (if all three parsers disagree).
-Candidates get weighted votes -- e.g., if parsers with weights $w_1 = 0.8$ and $w_2 = 0.7$ agree on the candidate, the candidate gets 1.5 votes.
-Since we have only three parsers, in practice this means that the candidate of the best parser looses only if 1.
-the other two parsers agree on someone else, or 2.
-if attaching the child to this candidate would create a cycle.
-
-The tree is constructed from the root down.
-We repeatedly add nodes whose winning parent candidates are already in the tree.
-If none of the remaining nodes meet this condition, we have to break a cycle.
-We do so by examining all unattached nodes.
-At each node we notice the votes of its current winning parent.
-Then we remove the least-scoring winner and go on with adding nodes until all nodes are attached or there is another cycle to break.
-
-\section{Experiments}
-\label{sec:experiments}
-
-The final test data are blind, any error analysis is therefore impossible.
-That is why all scores given in this section were measured on the development data.
-All three treebanks follow the same annotation scheme and each of them is available in two flavors:
+\section{What Is Interset and What It Is Not}
+\label{sec:whatisit}
 
 \begin{compactitem}
-\item \textit{nomorph} variety contains word forms, chunk labels, dependency links and dependency labels
-\item \textit{morph} variety is augmented by automatically assigned lemmas, part of speech tags and values of morphological features (gender, number, person, case, postposition and tam -- tense+aspect+modality)
+\item it is not a standard like eagles, parole, gold ontology and that sort of crap
+\item it is not a physical tagset (not intended to be used so - nemusim se ridit tim, co by radi lingvisti, aby jim to bylo bliz)
+\item it looses information but it never adds information. If source tagset does not encode something, it simply won't be in Interset.
+\item Specificaly, we do not retag words (say, we are converting from Brown to Penn, and hey, the word abc is tagged adverb in Brown and as conjunction in Penn...
+No!
+Unless we map all Brown adverbs on Interset conjunctions, which we probably don't want to, we will not change the tag of word abc.
+In fact, we NEVER look at the word the tag is assigned to.
+We work with the tag ONLY.
 \end{compactitem}
 
-\subsection{Morphology}
-\label{sec:morphology}
-
-\Tref{tab:baseline} shows baseline results on the \textit{nomorph} data. Both MST and Malt parsers were invoked in projective mode, Malt with the default Nivre arc-eager algorithm.
-
-\begin{table}[ht]
-\begin{centering}
-%\small
-\begin{tabular}{l|l|l|l|l}
-& \textbf{MST} & \textbf{Malt} & \textbf{DZ} & \textbf{Vote} \\
-\hline
-hi & 80.32 & 81.84 & 62.00 & \textbf{82.48}\\
-bn & 82.00 & \textbf{84.71} & 71.02 & 83.11\\
-te & 77.63 & \textbf{80.89} & 70.52 & 80.59\\
-\end{tabular}
-\caption{Baseline UAS of the four parsers on \textit{nomorph} development data. Language codes follow ISO 639: hi = Hindi, bn = Bangla, te = Telugu.}
-\label{tab:baseline}
-\end{centering}
-\end{table}
-
-There are several ways how to use the additional information from the \textit{morph} data.
-The easiest way, exploitable by all three parsers, is to combine the chunk label, the POS tag and the features into one tag string.
-The results (not presented here) are very poor.
-Although there may be tagging errors, the most likely cause is data sparseness.
-In \Tref{tab:corpus} we illustrate this by showing the numbers of unique values in the various attributes of treebank words.
-
-\begin{table}[ht]
-\begin{centering}
-%\small
-\begin{tabular}{l|r|r|r|r|r|r}
-& \textbf{occ} & \textbf{frm} & \textbf{lem} & \textbf{cl} & \textbf{pos} & \textbf{feat} \\
-\hline
-hi & 13779 & 3973 & 3134 & 10 & 33 & 714\\
-bn & 6449 & 2997 & 2336 & 14 & 30 & 367\\
-te & 5494 & 2462 & 1403 & 12 & 31 & 453\\
-\end{tabular}
-\caption{Size of the training corpora: occ -- word occurrences, frm -- distinct forms, lem -- lemmas, cl -- chunk labels, pos -- part of speech tags, feat -- feature value combinations.}
-\label{tab:corpus}
-\end{centering}
-\end{table}
-
-To fight sparseness, we could either restrict the tag to selected information, or split the information into multiple features learnt separately, or both.
-We first restricted the tag string to selected information.
-Parsing on other treebanks showed that POS with case is especially useful \citep{dzparser}.
-The other feature we selected is called \textit{vibhakti}, which partially corresponds to case suffix and partially to postposition.\footnote{The Indian treebanks at hand are unusual in that nodes do not always map to words.
-They represent chunks, with function words such as postpositions hidden in node attributes.}
-
-\Tref{tab:poscasevib} presents the results of restricting the tag string to POS+case+vibhakti.
-This move especially improves the MST Parser, which now outperforms Malt on Hindi and Bangla.
-Malt improves on Hindi but drops behind on Bangla and Telugu.
-DZ Parser also improves on Hindi and deteriorates elsewhere but there is an interesting observation: even though its own scores are worse, the worsened output actually improves the voting results (provided the configurations of MST and Malt are fixed).
-So it seems that the newly introduced errors are less important (because taken care of by the more powerful parsers) while some difficult parts of the data are now covered better.
-
-\begin{table}[ht]
-\begin{centering}
-%\small
-\begin{tabular}{l|l|l|l|l}
-& \textbf{MST} & \textbf{Malt} & \textbf{DZ} & \textbf{Vote}\\
-\hline
-hi & 86.16 & 82.96 & 75.12 & \XXX 86.64\\
-bn & 85.70 & 80.02 & 54.38 & \XXX 78.79\\
-te & 79.85 & 78.37 & 45.78 & \XXX 78.81\\
-\end{tabular}
-\caption{UAS on refined \textit{morph} data (POS tag, case and vibhakti concatenated). \XXX voting se pocitalo, kdyz jeste MST pracoval jen s nomorph, takze to musime prepocitat}
-\label{tab:poscasevib}
-\end{centering}
-% Když se ke značce POS přilepí pád, ale ne vibhakti, a s tím se přetrénuje Malt parser (ostatní parsery jsme nechali beze změny, tj.
-MST byl natrénovaný na nomorph a DZ naopak i na vibhakti), tak Malt sám má na všech třech jazycích pořád stejné výsledky (jako v této tabulce), ale výsledky Vote se na hindštině zhorší a na zbývajících dvou jazycích zlepší: hi=85.36, bn=83.35, te=80.89.
-\end{table}
-
-Finally, ve returned to \textit{nomorph} with Malt on Bangla and Telugu, where the POS+case+vibhakti tags did not help.
-The results are given in \Tref{tab:posmix2}.
-This is also the configuration we used to parse the test set for the official evaluation round 1.
-
-\begin{table}[ht]
-\begin{centering}
-%\small
-\begin{tabular}{l|l|l|l|l}
-& \textbf{MST} & \textbf{Malt} & \textbf{DZ} & \textbf{Vote}\\
-\hline
-hi & 86.16 & 85.84 & 75.12 & \textbf{87.12}\\
-bn & 85.70 & 84.71 & 54.38 & \textbf{86.19}\\
-te & 79.85 & 80.89 & 45.78 & \textbf{82.37}\\
-\end{tabular}
-\caption{UAS on mixed data: MST and DZ use POS+case+vibhakti for all languages, Malt uses that for Hindi and POS only elsewhere.
-MST is now the best parser for hi and bn, which changes the voting weights.}
-\label{tab:posmix2}
-\end{centering}
-\end{table}
-
-%\begin{compactitem}
-%\item \hi{स्टैंडर्डज} \textit{(sṭaiṁḍarḍaja)} \translit{(sṭaiṁḍarḍaja)}
-%\item \hi{स्टैंडर्डस} \textit{(sṭaiṁḍarḍasa)}
-%\item \hi{स्टैंडर्ड्स} \textit{(sṭaiṁḍarḍsa)}
-%\item \bn{সরিয়ে} Tohle je bengálsky.
-%\item \te{వాడుతున్నాం} Tohle je telugsky.
-%\end{compactitem}
-
-\subsection{Nonprojectivity}
-\label{sec:nonprojectivity}
-
-Nonprojectivity is a property of the dependency structure and the word order \citep{neproj} that makes parsing more difficult.
-All three parsers can produce nonprojective structures and all three treebanks are nonprojective.
-However, except for Hindi, the proportion of nonprojective dependencies is so small that one can hardly imagine that running the parsers in nonprojective mode would bring any improvement.
-A quick experiment with Malt Parser switched to the nonprojective Stack Eager algorithm revealed that it actually hurts the results even for Hindi.
-
-\begin{table}[ht]
-\begin{centering}
-%\small
-\begin{tabular}{l|ll}
-& \textbf{Edges} & \textbf{Sentences} \\
-\hline
-hi & 01.83 & 13.93\\
-bn & 00.96 & 05.49\\
-te & 00.45 & 01.31\\
-\end{tabular}
-\caption{Percentage of nonprojective dependencies, and of sentences containing at least one nonprojectivity.}
-\label{tab:nonprojectivity}
-\end{centering}
-\end{table}
-
-\subsection{Error Patterns}
-\label{sec:errors}
-
-The accuracy of the dependencies is relatively high and it is difficult to trace repetitive error patterns.
-In Hindi, many wrong attachments seem to be long-distance, and verbs, conjunctions, root and NULL nodes are frequently involved.
-Frequent words should perhaps be available to the parsers as parts of tag strings: for instance, Hindi \hi{कि} \translit{ki} ``that'' or \hi{तो} \translit{to} are wrongly attached because the parser only sees the general CC tag.
-On a similar note, problems with coordination, also observed e.g.
-by \citet{dzparser}, occur here, too: \hi{भाई और भाभी} \translit{bhāī aura bhābhī} ``brother and his wife'' is correctly recognized as coordination rooted by the conjunction \hi{और}, however, the conjunction node lacks the information about its noun children and fails to attach as the subject of the verb.
-
-The tag string should contain both the chunk label and the POS tag.
-So far we wrongly assumed that POS always determins the chunk label.
-It is often so but not always, as exemplified in the Bangla chunk sequence \bn{তবে সুদীপ ওকে একদিন আড়ালে ডেকে বলেছিল কৌতূহল দেখালে তুমি উঁচুতে উঠতে অনিমেষ} \translit{tabe sudīpa oke ekadina āṛāle ḍeke balechila kautūhala dekhāle tumi um̃cute uṭhate animeša}.
-The words \bn{ডেকে} and \bn{দেখালে} are tagged VGNF|VM while \bn{বলেছিল} and \bn{উঠতে} are VGF|VM.
-The parser gets them wrong and it could be caused by it seeing only VM in the tag.
-
-In Telugu, extraordinal number of sentences follow the SOV order so strongly that the last node (verb) is almost always attached to the root and most other nodes are attached directly to the last node.
-An example chunk sequence where this rule would lead to 100~\% accuracy follows: \te{రాష్ట్రంలొ రంగారెడ్డి మెదక్ నిజామాబాద్ జిల్లాలలొ పంటను గొప్పొ పండిస్తున్నారు} \translit{rāšṭraṁlo raṁgāreḍḍi medak nijāmābād jillālalo paṁṭanu goppo paṁḍistunnāru}.
-In the light of such examples it seems reasonable to provide the parsers with an additional feature telling whether a particular dependency observes the ``naïve Telugu'' structure.
-Note however that this will not help with the other two languages.
-While 73.75~\% of Telugu dependncies follow this rule, it is only 39.52~\% in Bangla and 35.71~\% in Hindi.
-
-\subsection{\XXX To Do}
-\begin{compactitem}
-\item \XXX Parser unique errors (oracle accuracy)
-\item \XXX How many times a candidate lost due to cycle prevention, and how many times this introduced an error?
-\item \XXX Learning curve
-\item \XXX Spočítat, kolik je ve kterém jazyce uzlů NULL.
-\item \XXX znacka chunku, priklad z bengalstiny
-\item \XXX casta slova, priklad z hindstiny
-\item \XXX koordinace, priklad z hindstiny
-\end{compactitem}
-
-\section{Official Evaluation}
-\label{sec:evaluation}
-
-Finally, we present the official evaluation of our voting superparser, as measured by the organizers on the test data.
-For this purpose, the parsing system has been retrained on both the training data and the development data.
-The results are shown in \Tref{tab:evaluation}.
-
-\begin{table}[ht]
-\begin{centering}
-\small
-\begin{tabular}{l|l|l|l}
-& \textbf{UAS} & \textbf{LAA} & \textbf{LAS} \\
-\hline
-hi & 88.58 (3:90.31) & 72.66 (4:76.38) & 68.60 (4:74.48)\\
-bn & 86.06 (4:90.32) & 71.28 (4:81.27) & 66.70 (5:79.81)\\
-te & 80.27 (4:86.28) & 54.20 (4:61.58) & 49.91 (4:60.55)\\
-\end{tabular}
-\caption{Official scores on the test data: unlabeled attachment score (UAS), label assignment accuracy (LAA) and labeled attachment score (LAS).
-The numbers in parentheses are the rank of our system and the score of the best system w.r.t.
-the given metric.}
-\label{tab:evaluation}
-\end{centering}
-\end{table}
-
-\section{Related and Future Work}
-\label{sec:related}
-
-There is a large body of work on parser combination.
-A summary can be found in \citet{nivre-mcdonald:2008:ACLMain}, whose approach is also related to ours w.r.t.
-the selection of parsers.
-However, their feature-based integration of MST and Malt parsers is much more sophisticated than our lightweight voting.
-Further improvement of accuracy can be expected if MST-Malt integration is applied to the Indian treebanks.
-
-Future work, at the time of writing, includes experiments that can be run before the evaluation round 2, and thus their results will appear in the final version of this paper.
-We definitely intend to test other algorithms of the Malt parser, as well as to reconfigure its feature pool and let it work with POS, case and vibhakti separately.
-
-Labeling of the dependencies is another problem that deserves more attention.
-We have concentrated on the unlabeled attachment score so far and for the sake of the official evaluation, we simply pushed the MST labels through.
-A separate postprocessing classifier would probably produce better results.
-
 \section{Conclusion}
-\label{sec:concl}
-
-We have described our system of voting parsers, as applied to the ICON 2009 NLP Tools Contest task.
-We showed that case and vibhakti are important features at least for parsing Hindi while their usability in Bangla and Telugu is limited by data sparseness.
-We also discussed several error patterns that could lead to further improvements of the parsing system in future.
+\label{sec:conclusion}
 
@@
 
-We are enormously grateful to the developers of MST and Malt parsers for making their software available to the research community.
-The research has been supported by the grant 
-MSM0021620838 (Czech Ministry of Education).
+%The research has been supported by the grant 
+%MSM0021620838 (Czech Ministry of Education).
 

### 2009-10-31 doc/papers/2010-icgl-hongkong/paper.tex
Protože v Hong Kongu už udeřila půlnoc, pro jistotu jsem rychle odeslal to, co mám, přestože je to příšerně nedokončené. Odeslaný stav teď konzervuju i v SVN.

@@
 \usepackage{amsmath}
+\usepackage{listings} % zdrojáky programů
 \usepackage{color} %pro korektury
@@
 %\newfontinstance\hi[Script=Devanagari]{Lohit Hindi}
+\newfontinstance\arfont[Script=Arabic]{Code2000}
 \newfontinstance\hifont[Script=Devanagari]{Code2000}
@@
 \newfontinstance\tefont[Script=Telugu]{Code2000}
+\newfontinstance\zhfont{Code2000}
 \newfontinstance\translitfont{Gentium}
+\newcommand{\ar}[1]{{\arfont #1}}
 \newcommand{\hi}[1]{{\hifont #1}}
@@
 \newcommand{\te}[1]{{\tefont #1}}
+\newcommand{\zh}[1]{{\zhfont #1}}
 \newcommand{\translit}[1]{{\translitfont \textit{(#1)}}}
@@
 \thispagestyle{fancy}
+\lstset{language=Perl}
+\lstset{basicstyle=\ttfamily} % Pro zdrojový kód použít stejný font, jaký používám všude v \texttt{}.
 
@@
 Tagset conversion is difficult, and solutions tend to be tailored to a particular pair of tagsets.
-We propose a universal approach that makes the conversion tools reusable.
-\XXX We also provide an indirect evaluation in the context of a parsing task.
+We discuss Interset, a universal approach that makes the conversion tools reusable.
+While some morphosyntactic categories are clearly defined and easily ported from one tagset to another, there are also phenomena that are difficult to deal with because of overlapping concepts.
+In the present paper we focus on some of such problems, discuss their coverage in selected tagsets and propose solutions to unify the respective tagsets' approaches.
 \end{abstract}
@@
 
-The rest of the paper is organized as follows: In \XXX we describe Interset and how it works [\XXX including encoding algorithm] [\XXX including applications: Petr Pořízka, Saša Rosen, můj parsing].
+%The rest of the paper is organized as follows: In \XXX we describe Interset and how it works [\XXX including encoding algorithm] [\XXX including applications: Petr Pořízka, Saša Rosen, můj parsing].
 In \XXX we describe our universal set of features.
@@
 
-\section{What Is Interset and What It Is Not}
-\label{sec:whatisit}
+The rest of the paper is organized as follows: In \Sref{sec:interset} we describe Interset and how it works.
+In \Sref{sec:features} we describe our universal set of features.
+Then, \Sref{sec:hard} lists decisions that are difficult w.r.t.
+universality, and proposes solutions.
+Finally, we demonstrate the implications on real tagsets, and provide illustrative statistics.
+
+\section{Interset}
+\label{sec:interset}
+
+Interset is a universal set of features and their values.
+It shall be able to store all features that are usually encoded in tags.
+The role of this universal set \textit{(“Interset”)} is similar to the role of Interlingua in Interlingua-based machine translation \citep{interlingua} or the role of Unicode among character sets.
+The Interset serves as an intermediate step on the way from tagset A to tagset B.
+The interaction between the Interset and tagsets A and B, respectively, is described in \textit{tagset drivers}.
+Once the drivers have been implemented, we can do the two-way conversion A to B and B to A, plus the conversion between one of these tagsets and any other tagset that has been defined so far.
+
+% Tenhle odstavec se případně může vypustit.
+We are not likely to spare much effort during the initial phase, if compared to just writing a targeted A-to-B conversion procedure.
+Actually, covering two completely new tagsets requires more work and care: we should describe both encoding and decoding of each tagset, we may have to think about features that are present in neither of them, and we will probably want to be more careful about aspects that may not matter to our current application.
+However, the reusability of the resulting code should compensate for the effort more than adequately.
+Plus we provide some algorithms to make adding new tagsets easier, and it is also possible that the required tagset has been covered by someone else who is sharing the code on the web.
+
+\subsection{A New Standard?}
+\label{sec:notastandard}
+
+\textbf{Interset is not a new annotation standard.} There have been attempts to standardize morphosyntactic tagging and it is not Interset's mission to compete with them.
+Instead, the goal is to cover as many existing tagsets as possible whether they conform to a standard or not.
+The set of Interset features and values could of course be compared to those defined in standards.
+There have been several European projects concerning tagset standardization.
+The EAGLES project \citep{eagles, LeechWilson1999} produced a set of recommendations for tagsets.
+Output of the LE-PAROLE project \citep{parole} was a multilingual corpus of 14 European languages, morphosyntactically annotated according to a common core PAROLE tagset, extended with a set of language specific features.
+Another multilingual corpus with common tagset is MULTEXT \citep{multext} for six European languages (English, French, Spanish, German, Italian and Dutch), and later its spin-off MULTEXT-EAST \citep{multext-east} for 12 languages (English, Bulgarian, Czech, Estonian, Hungarian, Romanian, Slovene, and in later versions also Croatian, Lithuanian, Resian, Russian and Serbian); the tagsets used in MULTEXT corpora comply with EAGLES.
+Various EAGLES-compliant tagsets can be added to our system and their mutual similarity will probably make adding them all easier.
+Weakly related is also the Gold Ontology project \citep{gold-ontology} that defines various linguistic concepts, some of which serve as feature names and feature values in Interset.
+Similarly, morphosyntactic and other terms are included in IsoCat.\footnote{\url{http://www.isocat.org/}}
+
+\subsection{A New Tagset?}
+\label{sec:notatagset}
+
+\textbf{Interset is not primarily meant as a new \textit{physical} tagset for annotation.} Although it obviously could be used that way (possibly after compressing the feature values), it is better thought of as a set of concepts that physical tagsets can map to.
+Design of physical tagsets is often guided by various needs such as conforming to linguistic tradition and terminology of the given language, minimizing errors of automatic disambiguation etc.
+In contrast, the most important design constraint for Interset is the portability of information from one tagset to the others.
+If a feature value X, encoded in tagset A, is not defined in tagset B but it is still probable that users of tagset B would tag the same set of words with feature value Y, then it is desirable that the Interset algorithms are able to change X to Y, when converting from A to B.
+At the same time, converting to tagset C might require changing X to Z, converting to tagset D would keep X (because D has X, too) and converting to other tagsets would result in resetting X to empty value because there is no better choice available.
+
+Conversion via Interset often looses information but never adds new information.
+Interset may define feature value X but it just won't be set unless the source tagset defines it, too.
+Specifically, the conversion procedure does not retag words.
+For instance, the source tagset may define one tag \texttt{IN} for both prepositions and subordinating conjunctions.
+The target tagset may have separate tags for each of those categories.
+So will Interset sort out the words tagged \texttt{IN} into prepositions and conjunctions?
+No!
+In fact, the procedure \textit{never} looks at the word the tag is assigned to.
+It only works with the tag itself.
+
+\subsection{Features}
+\label{sec:features}
 
 \begin{compactitem}
-\item it is not a standard like eagles, parole, gold ontology and that sort of crap
-\item it is not a physical tagset (not intended to be used so - nemusim se ridit tim, co by radi lingvisti, aby jim to bylo bliz)
-\item it looses information but it never adds information. If source tagset does not encode something, it simply won't be in Interset.
-\item Specificaly, we do not retag words (say, we are converting from Brown to Penn, and hey, the word abc is tagged adverb in Brown and as conjunction in Penn...
-No!
-Unless we map all Brown adverbs on Interset conjunctions, which we probably don't want to, we will not change the tag of word abc.
-In fact, we NEVER look at the word the tag is assigned to.
-We work with the tag ONLY.
+\item \texttt{\textbf{pos} = noun|adj|num|verb|adv|prep| conj|part|int|punc}
+\item \texttt{\textbf{subpos} = prop|class|pdt|det|art|aux| cop|mod|ex|voc|post|circ|preppron| comprep|coor|sub|comp|emp|res|inf|vbp}
+\item \texttt{\textbf{prontype} = prs|rcp|int|rel|dem|neg| ind|tot}
+\item \texttt{\textbf{numtype} = card|ord|mult|frac|gen}
+\item \texttt{\textbf{numform} = word|digit|roman}
+\item \texttt{\textbf{numvalue} = 1|2|3}
+\item \texttt{\textbf{advtype} = man|loc|tim|deg|cau}
+\item \texttt{\textbf{punctype} = peri|qest|excl|quot|brck| comm|colo|semi|dash|symb|root}
+\item \texttt{\textbf{puncside} = ini|fin}
+\item \texttt{\textbf{synpos} = subst|attr|adv|pred}
+\item \texttt{\textbf{poss} = poss}
+\item \texttt{\textbf{reflex} = reflex}
+\item \texttt{\textbf{negativeness} = pos|neg}
+\item \texttt{\textbf{definiteness} = ind|def|red}
+\item \texttt{\textbf{foreign} = foreign}
+\item \texttt{\textbf{gender} = masc|fem|com|neut}
+\item \texttt{\textbf{possgender} = masc|fem|com|neut}
+\item \texttt{\textbf{animateness} = anim|nhum|inan}
+\item \texttt{\textbf{number} = sing|dual|plu}
+\item \texttt{\textbf{case} = nom|gen|dat|acc|voc|loc|ins| ill|ine|ela|all|ade|abl|par|tmp|ter| tra|ess|abe|com|cau|dis}
+\item \texttt{\textbf{prepcase} = npr|pre}
+\item \texttt{\textbf{degree} = pos|comp|sub|abs}
+\item \texttt{\textbf{person} = 1|2|3}
+\item \texttt{\textbf{politeness} = inf|pol}
+\item \texttt{\textbf{subcat} = intr|tran}
+\item \texttt{\textbf{verbform} = fin|inf|sup|part|trans| ger}
+\item \texttt{\textbf{mood} = ind|imp|cnd|sub|jus}
+\item \texttt{\textbf{tense} = past|pres|fut}
+\item \texttt{\textbf{subtense} = aor|imp|pqp}
+\item \texttt{\textbf{aspect} = imp|perf}
+\item \texttt{\textbf{voice} = act|pass}
+\item \texttt{\textbf{abbr} = abbr}
+\item \texttt{\textbf{hyph} = hyph}
+\item \texttt{\textbf{style} = arch|form|norm|coll}
+\item \texttt{\textbf{typo} = typo}
+\item \texttt{\textbf{variant} = short|long|0|1|2|3|4|5|6| 7|8|9}
+\item \texttt{\textbf{other} = }\textit{any other info}
+\item \texttt{\textbf{tagset} = }\textit{where does the other info come from?}
 \end{compactitem}
+
+%\XXX
+The only reason of saving really everything is that converting a tagset to itself should not lose information.
+For that purpose we use the “other” feature.
+It contains arbitrary information that does not fit in other features and distinguishes tags.
+Since the information is not understood by any other tagset, we need to know which tagset the value comes from.
+Thus the identifier of the tagset should be stored in the “tagset” feature.
+
+Except for “tagset” and “other”, there is a predefined list of possible values for each feature.
+Every feature also allows the empty value.
+While several feature-based tagsets distinguish between unknown values and irrelevant features, we do not find it wise in Interset.
+For instance, the fifth character in the PDT Czech tagset identifies grammatical case.
+Its normal values are 1 to 7.
+For parts of speech that do not have case (e.g.
+interjections) the fifth character is \texttt{-} (dash).
+Adjectives generally do have case, yet there are borrowed words without Czech case suffixes whose case value is unknown (\texttt{X}).
+An example is the tag \texttt{AAIPX-{}-{}-{}-1A-{}-{}-{}-} for “Buenos” in Buenos Aires.
+The benefit of making this distinction explicit in a tagset is unclear.
+What is clear, however, is that we must not reflect it in the universal feature set.
+Who can say that a feature will be irrelevant—given the context of the values of the other features—in any tagset whatsoever?
+It is quite easy to find features that are relevant in one tagset and not the other: e.g.
+Czech past participles distinguish gender, English don’t.
+
+\section{Tagset Drivers}
+\label{sec:drivers}
+
+While the Interset is merely an abstract definition, the real implementation lies in the tagset drivers.
+A driver is a code library responsible for decoding and encoding tags.
+Decoding is reading a string (tag) into an internal data structure, in accordance with the list of possible features and their values.
+Encoding works the other way around.
+
+The encoder obviously is the more difficult part.
+The decoder just reads and sorts the information, ideally not losing a single piece of it.
+If anything has to be discarded because it does not fit the target tagset, the discarding is encoder’s task.
+There are two main reasons why encoding is not easy:
+
+\begin{compactitem}
+\item The encoder should be prepared to all values of all features, regardless that some of them are unknown in the particular tagset.
+For instance, if number = dual and the tagset does not know dual, it is probably better to encode plural than just leave number unknown.
+\item Even if the target tagset knows features A and B, concrete value of A can restrict permitted values of B.
+Some combinations of feature values are not allowed.
+For instance, the Swedish Parole tagset allows “pos = noun \& gender = common | neuter”, and also “pos = pronoun \& gender = masculine | feminine | common | neuter”.
+If we are to encode “pos = noun \& gender = masculine”, we can either honor the part of speech, or the gender, but not both.
+\end{compactitem}
+
+Fortunately enough, unknown feature values / combinations can be dealt with automatically if the driver has the list of all possible tags.
+By decoding all tags on the list, we get feature values for every tag.
+We thus know all feature values permitted in the given tagset and we know all value combinations.
+We have defined an ordered list of back-off values for every Interset feature value.
+The back-off lists contain all other values of the feature, including the empty value, so it is guaranteed that we always find a value that is permitted.\footnote{The necessary condition is that the decoder only sets known feature values, which is desirable anyway.} Of course, the encoder can override the default back-off list if necessary.
+
+As for unknown feature combinations, there is a predefined total ordering of the features that defines their priority (this can be overridden, too).
+Since features are ordered, all value combinations can be stored in a trie structure.
+On selecting value of a higher-priority feature, the structure immediately reveals restricted value space for all lower-priority features.
+
+This back-off technique is implemented in a helper module.
+Any driver can call it and have the features adjusted to something the driver itself might produce during decoding.
+The encoder can then concentrate on the driver’s native feature combinations.
+Besides that, the helper module can also check a driver’s integrity by looking whether the decoder only sets known features and values, whether encode(decode(x)) = x etc.
+
+The whole thing is implemented in Perl.
+The drivers are Perl modules whose encode and decode functions can be called from other Perl programs, either to access the feature values, or to convert tagsets.
+The conversion script is very simple and looks like this:\footnote{Real conversion script would also have to deal with the format in which the tags are mixed with text in the corpus.
+This example merely assumes a list of tags, without the actual words and other annotation.}
+
+\begin{lstlisting}
+use tagset::cs::pdt;
+use tagset::en::penn;
+while(<>)
+{
+    print tagset::en::penn::encode
+        tagset::cs::pdt::decode $_, "\n";
+}
+\end{lstlisting}
+
+We have implemented and tested drivers for several tagsets of the CoNLL 2006 (Buchholz and Marsi 2006) and 2007 (Nivre et al.
+2007) shared task treebanks, for the Penn Treebank (Marcus et al.
+1993), the Prague Dependency Treebank (Böhmová et al.
+2003) and others, totaling 20 drivers.
+Those drivers are freely available on the web.\footnote{\url{http://xxx.com/}}
+We believe that the reusability will only be truly exploited if the drivers are shared in the community and we encourage everyone to contribute with drivers they need to write for themselves.
+
+\section{The Hard Problems}
+\label{sec:hard}
+
+Working with various tagsets, we identified several fields that were difficult to capture and unify.
+
+\begin{compactitem}
+\item Pronouns, determiners, articles, wh-adverbs.
+\item Numerals.
+\item Indefinite verb form, especially participles.
+% Možná už na začátku se zmínit o tom, jak přistupujeme k chybějícím kategoriím a že někdy kategorie schválně rozdělujeme hierarchicky.
+Pokud zbude místo, tak taky ukázat tu čínskou sadu značek a proč jsme z ní dokázali převzít pouze zlomek informace.
+\item Particles and exotic small word categories.
+\item Different tokenizations, tags spanning multiple or joint tokens with incompatible categories. %stazene tvary, arab
+\end{compactitem}
+
+Endemic word classes were one example.
+Whenever seen fit, we tried to roof them with some more common parts of speech, instead of introducing a new high-level class.
+We wanted to reduce the necessity of encoders’ taking care of parts of speech unknown in their home tagsets.
+Roofed word classes are usually distinguishable by one of the detailed-part-of-speech features.
+
+Determiners, predeterminers and articles are one group of word classes missing in a substantial number of tagsets.
+We chose adjectives to serve as the roof class here.
+To pick another set of examples, here is an overview of various sorts of particles found in our tagsets:
+
+\begin{compactitem}
+\item unclassified particle (Czech \texttt{TT}, English \texttt{RP}, Swedish \texttt{Q})
+\item interrogative particle (Arabic \texttt{FI} \ar{هل} \translit{hal}, Bulgarian \texttt{Tn} \textit{ли} \translit{li})
+\item affirmative particle (Bulgarian \texttt{Ta} \textit{да} \translit{da})
+\item negative particle (Arabic \texttt{FN} \ar{لا} \translit{lā}, Bulgarian \texttt{Tn} \textit{не} \translit{ne}, German \texttt{PTKNEG} \textit{nicht})
+\item response particle (German \texttt{PTKANT} \textit{ja} = “yes”, \textit{nein} = “no”, \textit{doch} = “yes”, \textit{danke} = “thank you”…)
+\item auxiliary particle (Bulgarian \texttt{Tx} \textit{да} \translit{da} = “to”, \textit{ще} \translit{šte} = “will”)
+\item modal particle (Bulgarian \texttt{Tm} \textit{май} \translit{maj} = “possibly”)
+\item verbal particle (Bulgarian \texttt{Tv} \textit{нека} \translit{neka} = “let”)
+\item emphasis particle (Bulgarian \texttt{Te} \textit{даже} \translit{daže} = “even”)
+\item gradable particle (Bulgarian \texttt{Tg} \textit{най} \translit{naj} = “most”)
+\item unique POS (Danish \texttt{U}, covering the words \textit{at} = infinitival “to”, \textit{som}, \textit{der})
+\item infinitive mark (German \texttt{PTKZU} \textit{zu}, Swedish \texttt{IM} \textit{att}, English \texttt{TO} \textit{to} – includes prepositional occurrences of \textit{to})
+\item separated verbal prefix (German \texttt{PTKVZ}, \textit{vor} in \textit{stellen Sie sich vor})
+\item adjectival particle (German \texttt{PTKA}, \textit{am} in \textit{am besten}, \textit{zu} in \textit{zu groß})
+\item existential \textit{there} in English (\texttt{EX})
+\item measure word, quantifier (Chinese \texttt{DM})
+\item genitive particle de in Chinese (\texttt{DE}) \zh{的} and \zh{得}
+\item Chinese particles \zh{了} \translit{le} (perfect), \zh{著} \translit{zhe}, \zh{起} \translit{qǐ}, \zh{過} \translit{guò} (\texttt{Di})
+\item Chinese particles \zh{了} \translit{le}, \zh{的} \translit{de}, \zh{來} \translit{lái} (\texttt{Ta})
+\item Chinese particles \zh{而已} \translit{'éryǐ}, \zh{沒有} \translit{méiyǒu}, \zh{也罷} \translit{yěba}, \zh{沒有} \translit{méiyǒu}, \zh{好了} \translit{hǎole} (\texttt{Tb})
+\item Chinese particles \zh{呢} \translit{ne}, \zh{吧} \translit{ba}, \zh{啊} \translit{'a}, \zh{囉} \translit{luō} (\texttt{Tc})
+\item Chinese particles \zh{嗎} \translit{ma}, \zh{否} \translit{fǒu} (\texttt{Td})
+\end{compactitem}
+
+As mentioned earlier, some tagsets consider participles forms of verbs, others classify them as adjectives; some tagsets make numerals special cases of adjectives, others have separate POS tags for cardinals, ordinals and various other numeral classes, yet others separate cardinal numbers and put the rest under other POSes.
+Differences in approaches taken by different tagsets might result in different feature values; for instance, we could decode verbform = “participle” without regard to whether pos = “verb” or pos = “adj”.
+Naturally it is desirable to decode the same thing into the same set of features each time.
+Although we could ban particular feature-value combinations in Interset, effectively forcing the driver authors to seek the permitted decoding, we prefer to leave it as a recommendation, since we do not want to predict, which feature combinations will never ever be needed to distinguish two different words.
+%\XXX The recommending guidelines (part of Interset documentation) are another output of our study.
+
+Probably the broadest source of problems is pronouns, determiners and various WH-words.
+Somewhere pronouns are only personal or possessive; somewhere there is a diversity of interrogative, relative, demonstrative, indefinite and negative pronouns.
+In the BulTreeBank \citep{bultreebank}, anything interrogative is a pronoun, although it could be considered numeral \textit{(how much?)} or adverb \textit{(where?
+when?
+how?)} elsewhere.
+Some tagsets address the variable syntactic behavior of pronouns (\textit{he} substitutes a noun, \textit{his} functions as an adjective).
+Some tagsets and languages do not have determiners but they have pronouns (demonstrative, indefinite) instead.
+All that lead us to remove pronouns and determiners as independent parts of speech.
+Instead, nouns, adjectives and adverbs have the feature “prontype” to distinguish the various types (personal, demonstrative, interrogative…) Empty value of this feature signals a normal noun (adjective, adverb).
+
+Note however, that any guidelines are only to ensure unified approach to different presentations of the same information.
+It does not apply to information that simply is not there.
+If cardinals were tagged as \textit{normal} adjectives (without sub-classing adjectives to numeral and others) they would remain so in Interset and also in the target tagset.
+We cannot add information, we only can lose it.
 
@@
 \label{sec:conclusion}
+
+%\XXX (Tohle je zatím opsáno z LRECu, náš závěr by měl znít trochu jinak.)
+We have proposed a method for tagset conversion that is reusable and, to a reasonable extent, universal.
+Our interlingua-inspired approach enables to interpret part-of-speech and morphological tags in a uniform way, and to convert information that is shared by two tagsets.
+Besides the obvious advantage of being able to use tools that expect a particular tagset, we also observed improvements in performance of a statistical parser.
 

### 2009-11-01 doc/papers/2010-icgl-hongkong/paper.tex
V této verzi nakonec odesláno k posouzení.

@@
 
-%The rest of the paper is organized as follows: In \XXX we describe Interset and how it works [\XXX including encoding algorithm] [\XXX including applications: Petr Pořízka, Saša Rosen, můj parsing].
-In \XXX we describe our universal set of features.
-Then, \XXX lists decisions that are difficult w.r.t.
-universality, and propose solutions.
-Finally, we demonstrate the implications on real tagsets, and provide illustrative statistics.
-
 The rest of the paper is organized as follows: In \Sref{sec:interset} we describe Interset and how it works.
 In \Sref{sec:features} we describe our universal set of features.
+\Sref{sec:drivers} gives some implementation details of Interset-based conversion.
 Then, \Sref{sec:hard} lists decisions that are difficult w.r.t.
-universality, and proposes solutions.
-Finally, we demonstrate the implications on real tagsets, and provide illustrative statistics.
+universality, demonstrates them on real tagsets, and proposes solutions.
 
@@
 It shall be able to store all features that are usually encoded in tags.
-The role of this universal set \textit{(“Interset”)} is similar to the role of Interlingua in Interlingua-based machine translation \citep{interlingua} or the role of Unicode among character sets.
+The role of this universal set is similar to the role of Interlingua in Interlingua-based machine translation \citep{interlingua} or the role of Unicode among character sets.
 The Interset serves as an intermediate step on the way from tagset A to tagset B.
@@
 Actually, covering two completely new tagsets requires more work and care: we should describe both encoding and decoding of each tagset, we may have to think about features that are present in neither of them, and we will probably want to be more careful about aspects that may not matter to our current application.
-However, the reusability of the resulting code should compensate for the effort more than adequately.
-Plus we provide some algorithms to make adding new tagsets easier, and it is also possible that the required tagset has been covered by someone else who is sharing the code on the web.
+However, the reusability of the resulting code should compensate for the effort more than adequately; it is even possible that the required tagset has been covered by someone else who is sharing the code on the web.
+
+Besides abstract concept definitions, Interset also comprises some real software---supporting procedures that make adding new tagsets easier.
+Thanks to the encoding algorithm implemented in the support library, developers adding new tagsets need not (not necessarily) consider all features that are irrelevant to the tagset being added (but which may become relevant during conversion to a particular other tagset).
 
@@
 
+Interset has been used in cross-language parser adaptation \citep{ZemanResnik2008}, and in MorphCon, a GUI program that brings tag conversion to corpus users \citep{morphcon}.
+Currently, it is being deployed as a means of unified access to morphosyntactic annotation for the users of the parallel multilingual corpus Intercorp.\footnote{\url{http://www.korpus.cz/intercorp/?req=doc:uvod}}
+
 \subsection{A New Tagset?}
@@
 
-\textbf{Interset is not primarily meant as a new \textit{physical} tagset for annotation.} Although it obviously could be used that way (possibly after compressing the feature values), it is better thought of as a set of concepts that physical tagsets can map to.
+\textbf{Interset is not primarily meant as a new \textit{physical} tagset for annotation.} Although it obviously could be used that way (possibly after compressing the feature values), it is better thought of as a set of concepts that physical tagsets map to.
 Design of physical tagsets is often guided by various needs such as conforming to linguistic tradition and terminology of the given language, minimizing errors of automatic disambiguation etc.
 In contrast, the most important design constraint for Interset is the portability of information from one tagset to the others.
-If a feature value X, encoded in tagset A, is not defined in tagset B but it is still probable that users of tagset B would tag the same set of words with feature value Y, then it is desirable that the Interset algorithms are able to change X to Y, when converting from A to B.
+If a feature value X, encoded in tagset A, is not available in tagset B but it is still probable that users of tagset B would tag the same set of words with feature value Y, then it is desirable that the Interset algorithms are able to change X to Y, when converting from A to B.
 At the same time, converting to tagset C might require changing X to Z, converting to tagset D would keep X (because D has X, too) and converting to other tagsets would result in resetting X to empty value because there is no better choice available.
@@
 The target tagset may have separate tags for each of those categories.
-So will Interset sort out the words tagged \texttt{IN} into prepositions and conjunctions?
+Will then Interset sort out the words tagged \texttt{IN} into prepositions and conjunctions?
 No!
@@
 \label{sec:features}
+
+The current version of Interset defines the following features and values. Besides the values listed, every feature is also allowed to have an empty value.
 
@@
 
-%\XXX
-The only reason of saving really everything is that converting a tagset to itself should not lose information.
-For that purpose we use the “other” feature.
-It contains arbitrary information that does not fit in other features and distinguishes tags.
-Since the information is not understood by any other tagset, we need to know which tagset the value comes from.
-Thus the identifier of the tagset should be stored in the “tagset” feature.
+The \texttt{other} feature can store arbitrary information that did not fit elsewhere.
+Typically, that sort of information is unique to the given tagset (not just unique to the language!), and goes too far from what tagsets usually encode.
+We do not believe that there will ever be another tagset encoding such information---if we did, we would update Interset to provide corresponding feature/value.
+However, it still makes sense to store the information so that a tagset can be converted to itself and no information is lost.
+Since the value of \texttt{other} is not understood by any other tagset, we need to know which tagset the value comes from.
+Thus the identifier of the tagset should be stored in the \texttt{tagset} feature.
 
-Except for “tagset” and “other”, there is a predefined list of possible values for each feature.
-Every feature also allows the empty value.
+%\XXX Tenhle odstavec asi vyhodit.
 While several feature-based tagsets distinguish between unknown values and irrelevant features, we do not find it wise in Interset.
@@
 
-We have implemented and tested drivers for several tagsets of the CoNLL 2006 (Buchholz and Marsi 2006) and 2007 (Nivre et al.
-2007) shared task treebanks, for the Penn Treebank (Marcus et al.
-1993), the Prague Dependency Treebank (Böhmová et al.
-2003) and others, totaling 20 drivers.
-Those drivers are freely available on the web.\footnote{\url{http://xxx.com/}}
+At the time of writing there are drivers for 20 tagsets of 10 languages (Arabic, Bulgarian, Chinese, Czech, Danish, English, German, Polish, Portuguese and Swedish).
+Those drivers are freely available on the web.\footnote{\url{http://xxxxx.xx/}}
 We believe that the reusability will only be truly exploited if the drivers are shared in the community and we encourage everyone to contribute with drivers they need to write for themselves.
@@
 
+%\XXX Pokud zbude místo, tak taky ukázat tu čínskou sadu značek a proč jsme z ní dokázali převzít pouze zlomek informace.
+
+\subsection{Pronouns, Determiners, Articles and Wh-Adverbs}
+\label{sec:pronouns}
+
+Most languages and tagsets have personal pronouns, i.e.
+words like ``I'', ``you'', ``he'', ``she'', ``it''.
+Various interrogative and relative function words (wh-words in English) are often also considered pronouns.
+In addition, grammars of some languages distinguish \textit{determiners} while others prefer to categorize the same thing as a sort of pronouns.
+According to the definition of EAGLES, \textbf{pronoun} is a function word that \textit{replaces} a noun phrase, while \textbf{determiner} is a function word that \textit{modifies} a noun phrase.
+As a result, proper EAGLES-pronouns behave like nouns and determiners behave like adjectives.
+Note that possessive pronouns (i.e.
+``my'', ``your'', ``his'', ``her'', ``its''), also found in many languages, are personal possessive determiners in the sense of the EAGLES definition.
+
+% Následující odstavec by se dal doplnit z wiki a podobný by se dal z wiki přepsat i pro číslovky.
+Když ale jenom odkomentuju tenhle kus bez doplňování, tak zcela naplním 8 stran, a to jsou momentálně prázdné Acknowledgements.
+Takže po odanonymizování by to referencema přeteklo na 9.
+stranu, což je sice povoleno, ale jen malinko, což by nebylo hezké.
+Tak snad raději teď půjdu spát a počkám, jestli to vůbec přijmou a nebudou to chtít třeba zkrátit na poster.
+Abych teď pracně nepřidal něco, co pak stejně budu muset vyhodit.
+
+%\XXX In the BulTreeBank \citep{bultreebank}, anything interrogative is a pronoun, although it could be considered numeral \textit{(how much?)} or adverb \textit{(where?
+when?
+how?)} elsewhere.
+Some tagsets address the variable syntactic behavior of pronouns (\textit{he} substitutes a noun, \textit{his} functions as an adjective).
+Some tagsets and languages do not have determiners but they have pronouns (demonstrative, indefinite) instead.
+All that lead us to remove pronouns and determiners as independent parts of speech.
+Instead, nouns, adjectives and adverbs have the feature “prontype” to distinguish the various types (personal, demonstrative, interrogative…) Empty value of this feature signals a normal noun (adjective, adverb).
+
+Because tagsets often disagree in what is pronoun, what is determiner etc., it is difficult to find a unifying approach.
+We decided to limit the number of the major parts of speech in order to minimize the cases where a word would end up with an empty part of speech.
+If there were a part of speech called \textit{determiner}, drivers of tagsets not having determiners would either have to check whether \texttt{\textbf{pos} = det} during encoding, or they would fall back into a residual word class.
+On the other hand, if we tag determiners as special cases of adjectives (which is what Interset does), such drivers will simply encode determiners as adjectives (provided they have adjectives—but these are much more common).
+
+We also followed this solution with pronouns because of the following reasons:
 \begin{compactitem}
-\item Pronouns, determiners, articles, wh-adverbs.
-\item Numerals.
-\item Indefinite verb form, especially participles.
-% Možná už na začátku se zmínit o tom, jak přistupujeme k chybějícím kategoriím a že někdy kategorie schválně rozdělujeme hierarchicky.
-Pokud zbude místo, tak taky ukázat tu čínskou sadu značek a proč jsme z ní dokázali převzít pouze zlomek informace.
-\item Particles and exotic small word categories.
-\item Different tokenizations, tags spanning multiple or joint tokens with incompatible categories. %stazene tvary, arab
+\item Although pronouns are found in most tagsets, there is much controversy about the precise extent of that category.
+\item Some tagsets allow for distinguishing between \textit{substantive} and \textit{attributive} pronouns (roughly those that behave like nouns and those that behave like adjectives).
+Assigning pronouns to nouns and adjectives respectively helps preserve that distinction.
+\item Some of the features that distinguish pronouns from real nouns and adjectives (interrogativeness, for instance) are found with adverbs and numerals as well and thus it makes sense to separate them.
 \end{compactitem}
 
-Endemic word classes were one example.
-Whenever seen fit, we tried to roof them with some more common parts of speech, instead of introducing a new high-level class.
+Pronouns are recognized by nonempty value of the \texttt{prontype} feature; default type is \texttt{prs}, which denotes personal pronouns under nouns, and possessive pronouns under adjectives.
+The drawback of this approach is that the encoding procedure of a driver that recognizes pronouns must define its own method of asking the question \textit{``Does the current feature-value structure correspond to a pronoun?''}
+
+\subsection{Numerals}
+\label{sec:numerals}
+
+Numerals form an analogy to the pronoun-determiner problem.
+Most tagsets recognize cardinal numbers as a separate category.
+However, the rest is less obvious.
+Some tagsets recognize ordinal numbers while others classify them as adjectives because of their syntactic behavior.
+An extreme case is the Danish Parole tagset where all numbers including cardinals are adjectives; however, they form a distinguished subclass of adjectives and can thus be recognized.
+Tagsets of Czech and other Slavic languages define complex systems of numerals, according to traditional local grammars: besides cardinals and ordinals, there are separate tags for fractions (``one fifth''), multiplications (``five times''),\footnote{Note that in these languages such numerals can be expressed in one token, which justifies devising a tag for them.} adverbial ordinals (``for the first time''), various types of generic ordinals (``twofold'', ``four kinds of''), interrogative or relative numerals (``how many'', ``what'',\footnote{Here, the expected answer is ``first'', ``second'' etc.
+The ``what'' corresponds to different Slavic word than in e.g.
+``What do you do for living?'' An example from Czech is \textit{kolikátý}.} ``how many times''), demonstrative numerals (``that many'', ``that many times''), indefinite numerals (``few'', ``much'') etc.
+
+For the sake of consistency, the solution for numerals should be parallel to that of pronouns.
+Cardinal numbers, as the most specific and most widely recognized category, should retain their independence.
+The rest will be split among nouns (fractions), adjectives (ordinals and some generics) and adverbs (multiplications and ordinal points in time).
+Non-empty \texttt{numtype} feature will distinguish them from real nouns, adjectives and adverbs.
+Parallelly, \texttt{prontype} could be set, too, for interrogative, demonstrative and indefinite numerals.
+
+%\subsection{Counted Nouns in Slavic Languages}
+%\XXX Co s počítanými podstatnými jmény ve slovanských jazycích? To je ještě samostatný problém.
+
+\subsection{Non-Finite Verb Forms}
+\label{sec:participles}
+
+Non-finite verb forms often constitute mixed categories from the syntactic point of view.
+The syntactic properties of participles overlap with adjectives.
+Similarly, gerunds resemble nouns and transgressives function as adverbs.
+At the same time however, they retain their verbal arguments.
+
+Usually, these words are tagged as forms of verbs.
+However, there are exceptions.
+In the Swedish Parole tagset, participles are tagged as adjectives.
+Czech equivalent of gerunds is considered a deverbative noun; Czech participles have long and short forms, the former being tagged as adjectives, the latter as verbs.
+Some tagsets could even delimit participles as a separate part of speech, without clearly marking their affinity to adjectives or verbs.
+The Interset guidelines prefer marking all three categories (participles, gerunds and transgressives) as verbs.
+Note however that any guidelines are only to ensure unified approach to different presentations of the same information.
+It does not apply to information that simply is not there.
+If participles were tagged as \textit{normal} adjectives without sub-dividing adjectives into ``normal ones'' and participles, they would remain so in Interset and also in the target tagset.
+
+\subsection{Particles and Other Minor Parts of Speech}
+\label{sec:particles}
+
+Every tagset contains a number of specialized tags for small groups of words.
+They encode endemic phenomena that do not occur in other corpora/languages or simply are not distinguished there (because they are covered by other, broader-scope tags).
+Sometimes such tags are grouped as \textit{particles} or various ``residual'' classes.
+The approach of Interset is to roof them with some more common parts of speech, instead of introducing a new high-level class.
 We wanted to reduce the necessity of encoders’ taking care of parts of speech unknown in their home tagsets.
@@
 
-Determiners, predeterminers and articles are one group of word classes missing in a substantial number of tagsets.
-We chose adjectives to serve as the roof class here.
-To pick another set of examples, here is an overview of various sorts of particles found in our tagsets:
+An overview of various sorts of particles follows:
 
@@
 
-As mentioned earlier, some tagsets consider participles forms of verbs, others classify them as adjectives; some tagsets make numerals special cases of adjectives, others have separate POS tags for cardinals, ordinals and various other numeral classes, yet others separate cardinal numbers and put the rest under other POSes.
-Differences in approaches taken by different tagsets might result in different feature values; for instance, we could decode verbform = “participle” without regard to whether pos = “verb” or pos = “adj”.
-Naturally it is desirable to decode the same thing into the same set of features each time.
-Although we could ban particular feature-value combinations in Interset, effectively forcing the driver authors to seek the permitted decoding, we prefer to leave it as a recommendation, since we do not want to predict, which feature combinations will never ever be needed to distinguish two different words.
-%\XXX The recommending guidelines (part of Interset documentation) are another output of our study.
+\subsection{Alternate Values}
+\label{sec:valuearrays}
 
-Probably the broadest source of problems is pronouns, determiners and various WH-words.
-Somewhere pronouns are only personal or possessive; somewhere there is a diversity of interrogative, relative, demonstrative, indefinite and negative pronouns.
-In the BulTreeBank \citep{bultreebank}, anything interrogative is a pronoun, although it could be considered numeral \textit{(how much?)} or adverb \textit{(where?
-when?
-how?)} elsewhere.
-Some tagsets address the variable syntactic behavior of pronouns (\textit{he} substitutes a noun, \textit{his} functions as an adjective).
-Some tagsets and languages do not have determiners but they have pronouns (demonstrative, indefinite) instead.
-All that lead us to remove pronouns and determiners as independent parts of speech.
-Instead, nouns, adjectives and adverbs have the feature “prontype” to distinguish the various types (personal, demonstrative, interrogative…) Empty value of this feature signals a normal noun (adjective, adverb).
+Some tagsets contain tags that encode two or more values of the same feature at the same time.
+For instance, the Czech PDT tags define 11 values for the third character of the tag, which denotes gender and animateness.
+Four are more or less independent values: \texttt{M} = ``masculine animate'', \texttt{I} = ``masculine inanimate'', \texttt{F} = ``feminine'' and \texttt{N} = ``neuter''.
+The rest are various combinations: \texttt{Y=(M|I)}, \texttt{T=(I|F)}, \texttt{W=(I|N)}, \texttt{H} and \texttt{Q=(F|N)}, \texttt{Z=(M|I|N)}, \texttt{X=(M|I|F|N)}.
+The obvious goal here is to make life easier for taggers because some word forms are systematically identical for two or more genders.
 
-Note however, that any guidelines are only to ensure unified approach to different presentations of the same information.
-It does not apply to information that simply is not there.
-If cardinals were tagged as \textit{normal} adjectives (without sub-classing adjectives to numeral and others) they would remain so in Interset and also in the target tagset.
-We cannot add information, we only can lose it.
+Interset does not define separate feature values for all combinations.
+However, it does allow for storing alternate values if necessary.
+Since the feature structure is implemented as a Perl hash, a reference to an array of values can be assigned to any feature in the decoder:
+
+\begin{lstlisting}
+elsif($gender eq "H") 
+{ 
+    $f{gender} = ["fem", "neut"]; 
+}
+\end{lstlisting}
+
+At the first glance, such option poses a serious obstacle for encoder design.
+Before, the encoder was a nice sequence of \texttt{if} statements, merely saying ``if gender is \texttt{fem}, output character \texttt{F}''.
+Do we now need to check whether gender is array first?
+Even if we are writing driver for a tagset that never works with alternate values?
+Fortunately, no.
+Once again, the support library provides a procedure that can convert arrays to single values in all features except those that we acutally expect to contain arrays.
+
+There is nonetheless one open question about alternate values.
+Interset cannot express permissible combinations of multiple features.
+In the Czech example above, \texttt{Q} actually means more than just ``feminine or neuter''.
+It means ``feminine singular or neuter plural''.
+Interset can store both genders and both numbers but by that it will also cover feminine plurals and neuter singulars.
+
+\subsection{Joint Categories}
+\label{sec:joint}
+
+Every tagset assumes a certain tokenization of the tagged text.
+The tokenization guidelines may leave unsplit tokens that evolved historically from two words with separate categories.
+Examples include German \textit{zum = zu dem} ``to the'', Czech \textit{proň = pro něj} ``for him'' or \textit{žes = že jsi} ``that you have''.
+So in the German example, we have a token corresponding to two other tokens also occurring in the corpus, one tagged as preposition and the other as article.
+
+Much more difficult is Arabic where orthographic rules dictate not to space-separate certain sequences of words.
+So for instance, the conjunction \ar{و} \translit{wa} ``and'', often thrown in at the sentence beginnings, is connected to the following word, as are prepositions.
+Tokenization cannot be finished before morphological analysis because word segmentation may have ambiguous solutions.
+Thus, in the following example, we'll end up with long tags \texttt{CONJ+PREP+NOUN\_PROP} and \texttt{NOUN+CASE\_DEF\_NOM}, respectively.
+
+\lstset{language=XML}
+\begin{lstlisting}[escapechar=\%]
+<token_Arabic>%\ar{وبالفالوجة}%
+  <voc>wabiAlfAlwjp</voc>
+  <pos>wa/CONJ + bi/PREP +
+       AlfAlwjp/NOUN_PROP</pos>
+</token_Arabic>
+<token_Arabic>%\ar{مثال}%
+  <voc>mivAlu</voc>
+  <pos>mivAl/NOUN + u/CASE_DEF_NOM</pos>
+</token_Arabic>
+\end{lstlisting}
+
+Note that the first of the two examples corresponds to a sequence of three words (at least from the English perspective) while the second example describes just one noun (stem + suffix).
+The latter could be perfectly described in Interset; the former is the problem.
+
+Interset currently covers Arabic tagset of the Prague Arabic Dependency Treebank \citep{padt}, which is already tokenized.
+For the more moderate examples from Czech and German, it pretends that one of the joint categories is a special feature of the other.
+Thus for \textit{zum}, Interset would note that it's a preposition with a special property \textit{attached article}.
+It would not express any relation to how independent articles are tagged.
+Alternatively, we could use the array approach described in \Sref{sec:valuearrays} and assign two values to the \texttt{pos} feature.
+Neither solution is optimal because both are far from universal.
+In theory any part of speech could combine with any other and there could be more than two concatenated tokens, each with its own features that should not be mixed all together.
+Future versions of Interset might introduce a concept of arrays of sequential features structures (to solve the Arabic puzzle above) and arrays of parallel feature structures (to cover the \textit{permissible feature combinations} discussed in \Sref{sec:valuearrays}).
+In the meantime, the conversion program using Interset has to split joint tags using its own means.
+
+%\XXX Nějaká tabulka se statistikami?
 
@@
 
-%\XXX (Tohle je zatím opsáno z LRECu, náš závěr by měl znít trochu jinak.)
-We have proposed a method for tagset conversion that is reusable and, to a reasonable extent, universal.
-Our interlingua-inspired approach enables to interpret part-of-speech and morphological tags in a uniform way, and to convert information that is shared by two tagsets.
-Besides the obvious advantage of being able to use tools that expect a particular tagset, we also observed improvements in performance of a statistical parser.
+We have described Interset, a reusable and universal method for tagset conversion via common set of features and their values.
+We have shown how tags from a source tagset are \textit{decoded} into this intermediate feature space, and how their \textit{encoding} into target tagset can automatically replace unavailable feature values by carefully selected defaults that minimize possible damage.
+We also briefly compared Interset to tagset standardization efforts and pointed out the differences in goals between standards and Interset.
+Finally, we presented numerous examples from real tagsets to illustrate the most difficult parts of tagset conversion.
+The solutions we proposed pursue the ultimate goal that information loss is minimized and similar information is encoded similarly, across tagsets and languages.
 

### 2009-12-15 doc/papers/2010-icgl-hongkong/paper.tex
Konečná verze článku na ICGL. Zkráceno na 4 strany + reference, protože z toho udělali poster.

--- 2009-10-30 doc/papers/2010-icgl-hongkong/paper.tex
@@
 
-\title{Hard Problems of Tagset Conversion%
-% Tohle nechat zakomentované, je to jen tahák, jak udělat acknowledgement grantu při nedostatku místa. Jinak ale mám momentálně na konci opravdovou sekci Acknowledgements.
-%\thanks{ \hspace{.6em}The research has been supported by the grant 
-%MSM0021620838 (Czech Ministry of Education).}
+% Tahák, jak udělat acknowledgement grantu při nedostatku místa.
+\title{Hard Problems of Tagset Conversion\thanks{ \hspace{.6em}The research has been supported by the grant 
+MSM0021620838 (Czech Ministry of Education).}
 }
@@
 % Až to přestane být anonymní, tak také odkomentovat Acknowledgements a autoreference XXXX.
-\author{%Daniel Zeman\\
-%Univerzita Karlova v Praze, Ústav formální a aplikované lingvistiky\\
-%Malostranské náměstí 25, CZ-11800, Praha, Czechia\\ 
-%\texttt{zeman@ufal.mff.cuni.cz}
+\author{Daniel Zeman\\
+Univerzita Karlova v Praze, Ústav formální a aplikované lingvistiky\\
+Malostranské náměstí 25, CZ-11800, Praha, Czechia\\ 
+\texttt{zeman@ufal.mff.cuni.cz}
 }
@@
 
-Similarly, various NLP tools may depend on particular tagsets.
-While some tools indeed treat tags as atomic strings, others could exploit the tag structure to dig more information about the word—no matter whether they use the features in machine learning, or in human-designed rules.
-If the tagset changes, manual rules become useless and statistical models have to be retrained at least; even that may not be possible in case the training procedure works with selected subsets of the feature pool.
-Applicability of NLP software to multiple corpora is exactly the reason why one would want to convert tags from one tagset to another.
-
-For many tagset pairs, designing the conversion procedure is not easy.
-On one hand, there are rare tagsets (e.g.
-MULTEXT-EAST, \citet{multext-east}) fitting at the same time languages as distant as Czech and Estonian; on the other hand, tagsets of two closely related languages (e.g.
-Danish and Swedish) or even two tagsets of the same language may differ substantially (for instance, the Mamba tagset of Swedish \citep{talbanken} contains detailed classification of auxiliary verbs and punctuation but lacks features like number, mood, tense etc.; this is in sharp contrast to another Swedish tagset, Parole \citep{parolesv}, which in turn is not compatible with the Danish Parole \citep{paroleda} tagset (the former classifies participles as verb forms, the latter as adjective forms; the former has separate tags for numerals, the latter classifies both cardinal and ordinal numbers as adjectives; etc.)
+For many tagset pairs, designing the conversion procedure is not easy. On one hand, there are rare tagsets
+% (e.g. MULTEXT-EAST, \citet{multext-east})
+fitting at the same time languages as distant as Czech and Estonian; on the other hand, tagsets of two closely related languages (e.g.
+Danish and Swedish) or even two tagsets of the same language may differ substantially (for instance, the Mamba tagset of Swedish \citep{talbanken} contains detailed classification of auxiliary verbs and punctuation but lacks features like number, mood, tense etc.; this is in sharp contrast to another Swedish tagset, Parole \citep{parolesv}, which in turn is not compatible with the Danish Parole \citep{paroleda} tagset.
+% (the former classifies participles as verb forms, the latter as adjective forms; the former has separate tags for numerals, the latter classifies both cardinal and ordinal numbers as adjectives; etc.)
 
@@
 
-The rest of the paper is organized as follows: In \XXX we describe Interset and how it works [\XXX including encoding algorithm] [\XXX including applications: Petr Pořízka, Saša Rosen, můj parsing].
-In \XXX we describe our universal set of features.
-Then, \XXX lists decisions that are difficult w.r.t.
-universality, and propose solutions.
-Finally, we demonstrate the implications on real tagsets, and provide illustrative statistics.
+The rest of the paper is organized as follows: In \Sref{sec:interset} we describe Interset and how it works.
+Then, \Sref{sec:hard} lists decisions that are difficult w.r.t.
+universality, demonstrates them on real tagsets, and proposes solutions.
 
-\section{What Is Interset and What It Is Not}
-\label{sec:whatisit}
+\section{Interset}
+\label{sec:interset}
 
+Interset is a universal set of features and their values.
+It shall be able to store all features that are usually encoded in tags.
+The role of this universal set is similar to the role of Interlingua in Interlingua-based machine translation \citep{interlingua} or the role of Unicode among character sets.
+The Interset serves as an intermediate step on the way from tagset A to tagset B.
+The interaction between the Interset and tagsets A and B, respectively, is described in \textit{tagset drivers}.
+Once the drivers have been implemented, we can do the two-way conversion A to B and B to A, plus the conversion to/from any other tagset that has been defined so far.
+
+Besides abstract concept definitions, Interset also comprises some real software---supporting procedures that make adding new tagsets easier.
+Thanks to the encoding algorithm implemented in the support library, developers adding new tagsets need not (not necessarily) consider all features that are irrelevant to the tagset being added (but which may become relevant during conversion to a particular other tagset).
+
+At the time of writing there are drivers for 20 tagsets of 10 languages, freely available on-line.\footnote{\url{https://wiki.ufal.ms.mff.cuni.cz/user:zeman:interset}}
+
+%\XXX Vypustil jsem celou kapitolu o tom, jak fungují ovladače. Tím se ale ztratila informace o tom, že encoding řeší defaulty. Nevím, jestli se to tam nemá nějak stručně vrátit.
+
+\subsection{A New Standard?}
+\label{sec:notastandard}
+
+\textbf{Interset is not a new annotation standard.} There have been attempts to standardize morphosyntactic tagging and it is not Interset's mission to compete with them.
+Instead, the goal is to cover as many existing tagsets as possible whether they conform to a standard or not.
+The set of Interset features and values could of course be compared to those defined in standards.
+There have been several European projects concerning tagset standardization.
+The EAGLES project \citep{eagles, LeechWilson1999} produced a set of recommendations for tagsets.
+Output of the LE-PAROLE project \citep{parole} was a multilingual corpus of 14 European languages, morphosyntactically annotated according to a common core PAROLE tagset, extended with a set of language specific features.
+Another multilingual corpus with common tagset is MULTEXT \citep{multext} for six European languages (en, fr, es, de, it, nl), and later its spin-off MULTEXT-EAST \citep{multext-east} for 12 languages (en, bg, cs, ee, hu, ro, sl, later also hr, lt, ru, sl-res);
+%(English, Bulgarian, Czech, Estonian, Hungarian, Romanian, Slovene, later also Croatian, Lithuanian, Russian and Resian);
+the tagsets used in MULTEXT corpora comply with EAGLES.
+Various EAGLES-compliant tagsets can be added to our system and their mutual similarity will probably make adding them all easier.
+Weakly related is also the Gold Ontology project \citep{gold-ontology} that defines various linguistic concepts, some of which serve as feature names and feature values in Interset.
+Similarly, morphosyntactic and other terms are included in IsoCat.\footnote{\url{http://www.isocat.org/}}
+
+Interset has been used in cross-language parser adaptation \citep{ZemanResnik2008}, and in MorphCon, a GUI program that brings tag conversion to corpus users \citep{morphcon}.
+Currently, it is being deployed as a means of unified access to morphosyntactic annotation for the users of the parallel multilingual corpus Intercorp.\footnote{\url{http://www.korpus.cz/intercorp/?req=doc:uvod}}
+
+\subsection{A New Tagset?}
+\label{sec:notatagset}
+
+\textbf{Interset is not primarily meant as a new \textit{physical} tagset for annotation.} Although it obviously could be used that way (possibly after compressing the feature values), it is better thought of as a set of concepts that physical tagsets map to.
+%Design of physical tagsets is often guided by various needs such as conforming to linguistic tradition and terminology of the given language, minimizing errors of automatic disambiguation etc.
+Physical tagsets often need to conform to linguistic tradition and terminology of the given language, minimize tagging errors etc.
+In contrast, the most important design constraint for Interset is the portability of information from one tagset to the others.
+If a feature value X is known in tagset A and unknown in B, and users of B are likely to tag the same words with feature value Y, then the Interset algorithms should replace X by Y.
+%If a feature value X, encoded in tagset A, is not available in tagset B but it is still probable that users of tagset B would tag the same set of words with feature value Y, then it is desirable that the Interset algorithms are able to change X to Y, when converting from A to B.\footnote{See \citet{biblio:ZeReusableTagset2008} for details on the algorithm.}
+%At the same time, converting to tagset C might require changing X to Z, converting to tagset D would keep X (because D has X, too) and converting to other tagsets would result in resetting X to empty value because there is no better choice available.
+
+Conversion via Interset often looses information but never adds new information.
+Interset may define feature value X but it just won't be set unless the source tagset defines it, too.
+Specifically, the conversion procedure does not retag words.
+For instance, the source tagset may define one tag \texttt{IN} for both prepositions and subordinating conjunctions.
+The target tagset may have separate tags for each of those categories but Interset will not sort out the words tagged \texttt{IN}.
+In fact, the procedure \textit{never} looks at the word the tag is assigned to.
+It only works with the tag itself.
+
+\section{The Hard Problems}
+\label{sec:hard}
+
+\subsection{Pronouns, Determiners, Wh-Adverbs}
+\label{sec:pronouns}
+
+Most languages and tagsets have personal pronouns, i.e.
+words like ``I'', ``you'', ``he''.
+Various interrogative and relative function words are often also considered pronouns.
+In addition, grammars of some languages distinguish \textit{determiners} while others prefer to categorize the same thing as a sort of pronouns.
+According to the definition of EAGLES, \textbf{pronoun} is a function word that \textit{replaces} a noun phrase, while \textbf{determiner} is a function word that \textit{modifies} a noun phrase.
+As a result, proper EAGLES-pronouns behave like nouns and determiners behave like adjectives.
+Note that possessive pronouns (i.e.
+``my'', ``your''), also found in many languages, are personal possessive determiners in the sense of the EAGLES definition.
+
+Because tagsets often disagree in what is pronoun, what is determiner etc., it is difficult to find a unifying approach.
+We decided to limit the number of the major parts of speech in order to minimize the cases where a word would end up with an empty part of speech.
+If there was a part of speech called \textit{determiner}, drivers of tagsets not having determiners would either have to check whether \texttt{\textbf{pos} = det} during encoding, or they would fall back into a residual word class.
+On the other hand, if we tag determiners as special cases of adjectives (which is what Interset does), such drivers will simply encode determiners as adjectives (which are much more common).
+
+We also followed this solution with pronouns because of the following reasons:
 \begin{compactitem}
-\item it is not a standard like eagles, parole, gold ontology and that sort of crap
-\item it is not a physical tagset (not intended to be used so - nemusim se ridit tim, co by radi lingvisti, aby jim to bylo bliz)
-\item it looses information but it never adds information. If source tagset does not encode something, it simply won't be in Interset.
-\item Specificaly, we do not retag words (say, we are converting from Brown to Penn, and hey, the word abc is tagged adverb in Brown and as conjunction in Penn...
-No!
-Unless we map all Brown adverbs on Interset conjunctions, which we probably don't want to, we will not change the tag of word abc.
-In fact, we NEVER look at the word the tag is assigned to.
-We work with the tag ONLY.
+\item Although pronouns are found in most tagsets, there is much controversy about the precise extent of that category.
+\item Some tagsets allow for distinguishing between \textit{substantive} and \textit{attributive} pronouns. Assigning pronouns to nouns and adjectives respectively helps preserve that distinction.
+\item Some of the features that distinguish pronouns from real nouns and adjectives (interrogativeness, for instance) are found with adverbs and numerals as well and thus it makes sense to separate them.
 \end{compactitem}
+
+Pronouns are recognized by nonempty value of the \texttt{prontype} feature.
+%; default type is \texttt{prs}, which denotes personal pronouns under nouns, and possessive pronouns under adjectives. 
+The drawback of this approach is that the encoding procedure of a driver that recognizes pronouns must define its own method of asking the question \textit{``Does the current feature structure correspond to a pronoun?''}
+
+\subsection{Numerals}
+\label{sec:numerals}
+
+Numerals form an analogy to the pronoun-determiner problem.
+Most tagsets recognize cardinal numbers as a separate category.
+However, the rest is less obvious.
+Some tagsets recognize ordinal numbers while others classify them as adjectives because of their syntactic behavior.
+%An extreme case is the Danish Parole tagset where all numbers including cardinals are adjectives; however, they form a distinguished subclass of adjectives and can thus be recognized.
+%Tagsets of Czech and other Slavic languages define complex systems of numerals, according to traditional local grammars: besides cardinals and ordinals, there are separate tags for fractions (``one fifth''), multiplications (``five times''),\footnote{Note that in these languages such numerals can be expressed in one token, which justifies devising a tag for them.} adverbial ordinals (``for the first time''), various types of generic ordinals (``twofold'', ``four kinds of''), interrogative or relative numerals (``how many'', ``what'',\footnote{Here, the expected answer is ``first'', ``second'' etc.
+The ``what'' corresponds to different Slavic word than in e.g.
+``What do you do for living?'' An example from Czech is \textit{kolikátý}.} ``how many times''), demonstrative numerals (``that many'', ``that many times''), indefinite numerals (``few'', ``much'') etc.
+Some Slavic tagsets define complex systems of numerals, according to traditional local grammars: besides cardinals and ordinals, there are separate tags for fractions, multiplications, adverbial ordinals, generic ordinals, interrogative, relative, demonstrative or indefinite numerals.
+
+For the sake of consistency, the solution should be parallel to that of pronouns.
+Cardinal numbers, as the most specific and most widely recognized category, should retain their independence.
+The rest will be split among nouns (fractions), adjectives (ordinals and some generics) and adverbs (multiplications and ordinal points in time).
+Non-empty \texttt{numtype} feature will distinguish them from real nouns, adjectives and adverbs.
+Parallelly, \texttt{prontype} could be set, too, for interrogative, demonstrative and indefinite numerals.
+
+\subsection{Non-Finite Verb Forms}
+\label{sec:participles}
+
+Non-finite verb forms often constitute mixed categories from the syntactic point of view.
+The syntactic properties of participles overlap with adjectives.
+Similarly, gerunds resemble nouns and transgressives function as adverbs.
+At the same time however, they retain their verbal arguments.
+
+Usually, these words are tagged as forms of verbs.
+However, there are exceptions.
+In the Swedish Parole tagset, participles are tagged as adjectives.
+Czech equivalent of gerunds is considered a deverbative noun.
+%; Czech participles have long and short forms, the former being tagged as adjectives, the latter as verbs.
+Some tagsets could even delimit participles as a separate part of speech, without clearly marking their affinity to adjectives or verbs.
+The Interset guidelines prefer marking all three categories (participles, gerunds and transgressives) as verbs.\footnote{Note however that any guidelines are only to ensure unified approach to different presentations of the same information.
+% It does not apply to information that simply is not there. 
+If participles were tagged as \textit{normal} adjectives without sub-dividing adjectives into ``normal ones'' and participles, they would remain so in Interset and also in the target tagset.}
+
+\subsection{Alternate Values}
+\label{sec:valuearrays}
+
+Some tagsets contain tags that encode two or more values of the same feature at the same time.
+For instance, the Czech PDT tags define 11 values for the third character of the tag, which denotes gender and animateness.
+Four are more or less independent values: \texttt{M} = ``masculine animate'', \texttt{I} = ``masculine inanimate'', \texttt{F} = ``feminine'' and \texttt{N} = ``neuter''.
+The rest are various combinations: \texttt{Y=(M|I)}, \texttt{T=(I|F)}, \texttt{W=(I|N)}, \texttt{H} and \texttt{Q=(F|N)}, \texttt{Z=(M|I|N)}, \texttt{X=(M|I|F|N)}.
+The obvious goal here is to make life easier for taggers because some word forms are systematically ambiguous.
+
+Interset does not define separate feature values for all combinations. However, it does allow for storing alternate values if necessary:
+%. Since the feature structure is implemented as a Perl hash, a reference to an array of values can be assigned to any feature in the decoder:
+
+%elsif($gender eq "H") 
+%{ 
+\begin{lstlisting}
+$f{gender} = ["fem", "neut"]; 
+\end{lstlisting}
+%}
+
+At the first glance, such option poses a serious obstacle for encoder design.
+Before, the encoder was a nice sequence of \texttt{if} statements, merely saying ``if gender is \texttt{fem}, output character \texttt{F}''.
+Do we now need to check whether gender is array first?
+Even if we are writing driver for a tagset that never works with alternate values?
+Fortunately, no.
+Once again, the support library provides a procedure that can convert arrays to single values in all features except those that we acutally expect to contain arrays.
+
+There is nonetheless one open question about alternate values.
+Interset cannot express permissible combinations of multiple features.
+In the Czech example above, \texttt{Q} actually means more than just ``feminine or neuter''.
+It means ``feminine singular or neuter plural''.
+Interset can store both genders and both numbers but by that it will also cover feminine plurals and neuter singulars.
+
+\subsection{Joint Categories}
+\label{sec:joint}
+
+Every tagset assumes a certain tokenization of the tagged text.
+The tokenization guidelines may leave unsplit tokens that evolved historically from two words with separate categories.
+Examples include German \textit{zum = zu dem} ``to the'', Czech \textit{proň = pro něj} ``for him''.
+% or \textit{žes = že jsi} ``that you have''.
+So in the German example, we have a token corresponding to two other tokens also occurring in the corpus, one tagged as preposition and the other as article.
+We then have a token corresponding to two other tokens also occurring in the corpus, with different POS.
+
+Much more difficult is Arabic where orthographic rules dictate not to space-separate certain sequences of words.
+So for instance, the conjunction \ar{و} \translit{wa} ``and'', often thrown in at the sentence beginnings, is connected to the following word, as are prepositions.
+Tokenization cannot be finished before morphological analysis because word segmentation may have ambiguous solutions.
+Thus, in the following example, we'll end up with long tags \texttt{CONJ+PREP+NOUN\_PROP} and \texttt{NOUN+CASE\_DEF\_NOM}, respectively.
+
+\lstset{language=XML}
+\begin{lstlisting}[escapechar=\%]
+<token_Arabic>%\ar{وبالفالوجة}%
+  <pos>wa/CONJ + bi/PREP +
+       AlfAlwjp/NOUN_PROP</pos>
+<token_Arabic>%\ar{مثال}%
+  <pos>mivAl/NOUN + u/CASE_DEF_NOM</pos>
+\end{lstlisting}
+%  <voc>wabiAlfAlwjp</voc>
+%</token_Arabic>
+%  <voc>mivAlu</voc>
+%</token_Arabic>
+
+Note that the first of the two examples corresponds to a sequence of three words (at least from the English perspective) while the second example describes just one noun (stem + suffix).
+The latter could be perfectly described in Interset; the former is the problem.
+
+%Interset currently covers Arabic tagset of the Prague Arabic Dependency Treebank \citep{padt}, which is already tokenized.
+For the more moderate examples from Czech and German, it pretends that one of the joint categories is a special feature of the other.
+For the Czech and German examples above, Interset encodes one of the joint categories as a feature of the other.
+Thus for \textit{zum}, Interset would note that it's a preposition with a special property \textit{attached article}.
+It would not express any relation to how independent articles are tagged.
+Alternatively, we could use the array approach described in \Sref{sec:valuearrays} and assign two values to the \texttt{pos} feature.
+Neither solution is optimal because both are far from universal.
+In theory any part of speech could combine with any other and there could be more than two concatenated tokens, each with its own features that should not be mixed all together.
+%Future versions of Interset might introduce a concept of arrays of sequential features structures (to solve the Arabic puzzle above) and arrays of parallel feature structures (to cover the \textit{permissible feature combinations} discussed in \Sref{sec:valuearrays}).
+In the meantime, the conversion program using Interset has to split joint tags using its own means.
 
@@
 
-\section*{Acknowledgements}
+We have described Interset, a reusable and universal method for tagset conversion via common set of features and their values.
+We briefly compared Interset to tagset standardization efforts and pointed out the differences in goals between standards and Interset.
+Then we presented numerous examples from real tagsets to illustrate the most difficult parts of tagset conversion.
+The solutions we proposed pursue the ultimate goal that information loss is minimized and similar information is encoded similarly, across tagsets and languages.
+
+% Šetříme místem, granty uvádíme v poznámce na první straně.
+%\section*{Acknowledgements}
 
